Implicit motor adaptation patterns in a redundant motor task manipulating a stick with both hands

Version 1

Abstract

Abstract

The remarkable ability of the motor system to adapt to novel environments has traditionally been investigated using kinematically non-redundant tasks, such as planar reaching movements. Due to this limitation, it remains poorly understood how the motor system achieves adaptation by altering the movement patterns of our redundant body. To address this issue, we developed a redundant motor task in which participants reached for targets with the tip of a virtual stick held with both hands. Despite the redundancy of the task, participants consistently employed a stereotypical strategy of flexibly changing the tilt angle of the stick depending on the direction of tip movement. Thus, this baseline relationship between tip-movement direction and stick-tilt angle constrained both the physical and visual movement patterns of the redundant system. Our task allowed us to systematically investigate how the motor system implicitly changed both the tip-movement direction and the stick-tilt angle in response to imposed visual perturbations. Both types of perturbations, whether directly affecting the task (tip-movement direction) or not (stick-tilt angle around the tip), drove adaptation, and the patterns of implicit adaptation were guided by the baseline relationship. Consequently, tip-movement adaptation was associated with changes in stick-tilt angle, and intriguingly, even seemingly ignorable stick-tilt perturbations significantly influenced tip-movement adaptation, leading to tip-movement direction errors. These findings provide a new understanding that the baseline relationship plays a crucial role not only in how the motor system controls movement of the redundant system, but also in how it implicitly adapts to modify movement patterns.

Introduction

If a movement error occurs while reaching a target, the motor system automatically and implicitly updates the motor command for the next movement to reduce the error. This remarkable motor learning ability contributes not only to maintaining accurate movements in the presence of noise but also to flexibly adapting the movement to new environments. In the standard planar arm-reaching task, which has been widely used to study the mechanisms of motor learning, participants manipulate a visual cursor with their hands primarily by moving their shoulder and elbow joints (Flash & Hogan, 1985; Gordon et al., 1994) (Fig. 1a). The arm-reaching task has contributed to uncovering many important properties of motor adaptation when a novel environment is introduced (Thoroughman & Shadmehr, 1999; Mazzoni & Krakauer, 2006; Taylor & Ivry, 2011; Takiyama et al., 2015; Hayashi et al., 2016).

Kinematic redundancy and adaptation patterns.

(a) In the ordinary arm- reaching movement task, there is no redundancy between the task goal (xc, yc) and the joint angles (𝜃1, 𝜃2). (b) The movement pattern at each target is uniquely determined by the joint angles. (c) The adaptation patterns to a visual rotation are also uniquely determined. (d) This adapted joint angle pattern should be identical to that of voluntary reaching in the adapted direction. (e, f) In the case of planar reaching movement by a hypothetical limb with three joints (𝜃1, 𝜃2, 𝜃3), the cursor can reach the target in infinite patterns. (g) Given that the motor system solves the task redundancy by optimization, we should observe a stereotypical pattern of joint angles according to the targets. (h, i) The question is whether, when a visual rotation is imposed on the cursor (end-effector relevant perturbation), the motor system adopts the same joint movement pattern as when voluntarily aiming in the adapted direction (left) or a new joint movement pattern (right). (j) The joint angles can be perturbed while the cursor position remains unchanged (end-effector irrelevant perturbation). (k) This perturbation can be ignored because it does not affect the cursor (left). Alternatively, the joint angles may be corrected (right).

Notably, there is no redundancy between hand position and joint angles in this task. The joint angles are uniquely determined by the target to be reached (Fig. 1b). Because of this non-redundancy, when the motor system adapts the reaching movement to the visual rotation imposed on the cursor (Fig. 1c, left), the adapted pattern of joint angles is also uniquely determined (Fig. 1c, right). Naturally, the adapted pattern of joint angles should be identical to the pattern when reaching in the adapted direction (Fig. 1d). However, our body systems for performing practical motor tasks generally have more degrees of freedom than necessary to accomplish the tasks (Bernstein, 1996; Singh et al., 2016). For example, in the golf swing, there are countless patterns of whole-body configurations to achieve identical clubhead positions at ball contact.

The problem arising from such a redundant structure can be illustrated by considering the reaching movement of a hypothetical limb with three joints (Fig. 1e). In this case, there are an infinite number of possible joint angle patterns for the same hand position (Fig. 1f). However, given that the motor system solves task redundancy by minimizing motor cost (Collins, 1995; O’Sullivan et al., 2009) or movement error (White & Diedrichsen, 2010; Zhang et al., 2018), we should observe a stereotypical pattern of joint angles according to the targets (Fig. 1g). The problem is how the motor system changes the direction of cursor movement to implicitly adapt to the visual rotation imposed on the cursor (Fig. 1h). After adaptation, does the motor system move the joint angles in the same way that it voluntarily aims in the direction of the adapted hand movement? (Fig. 1i, left). Alternatively, does the motor system adopt a new joint movement pattern? (Fig. 1i, right). It has not been thoroughly investigated how the motor system coordinates to change the redundant body system to implicitly adapt to the perturbation of the end-effector (i.e., cursor or hand).

Redundancy poses another type of problem for implicit movement adaptation. If the joint angles are perturbed while the hand position remains unchanged (Fig. 1j), how does the motor system respond to the perturbation? The motor system may simply ignore the perturbation because it does not affect the cursor position (Fig. 1k, left). Alternatively, the motor system may attempt to correct the joint movements (Fig.1k, right). Conventionally, this type of problem has been investigated in the context of how the motor system responds to the movement variability that affects or does not affect task performance (i.e., task-relevant or task-irrelevant variability). Previous studies have consistently shown that the motor system actively corrects for task- relevant variability while allowing task-irrelevant variability (Mosier et al., 2005; Latash, 2012; van Beers et al., 2013; Sternad, 2018; Scholtz & Schoner, 1999; Scholtz et al., 2000). Optimal feedback control theory suggests that the strategies based on such a "minimal intervention principle" are optimal in terms of achieving task goals (e.g., movement accuracy) (Todorov & Jordan, 2002; Todorov, 2004; Scott, 2004; Diedrichsen, 2007; Valero-Cuevas et al., 2009; Ronsse et al., 2010; Sternad et al., 2011; Selgrade & Chang, 2015). These studies predict that the motor system should ignore the perturbations to joint angles that do not affect the task goal (Fig 1k, left). However, these studies cannot directly test whether the redundant body movement pattern is also unchanged when the end-effector movement remains unchanged.

Conversely, it has also been reported that the motor system adapts to change the movement in response to the task-irrelevant perturbation (Wolpert et al., 1995; Flanagan & Rao, 1995; Morehead et al., 2017; Diedrichsen et al., 2010; Schaefer et al., 2012; Kim et al., 2019). For example, when the cursor trajectory was curved only in the middle of the reaching movement, the hand movement was adaptively modified even though the perturbation did not cause a reaching error at the target (Wolpert et al., 1995; Flanagan & Rao, 1995). These studies created the task-irrelevant condition by applying the perturbation to the cursor (Wolpert et al., 1995; Flanagan & Rao, 1995; Morehead et al., 2017) or by combining the cursor perturbation with the target shape manipulation (Diedrichsen et al., 2010; Schaefer et al., 2012; Kim et al., 2019). Thus, the problem of motor adaptation to such task-irrelevant perturbations in the previous studies is totally different from the problem of adaptation arising from body redundancy (Figs. 1i and 1k). In light of this, in this study, we use the terms "end- effector relevant perturbation" (e.g., Fig. 1h) and "end-effector irrelevant perturbation" (e.g., Fig. 1j) when appropriate, instead of the more commonly used terms "task- relevant perturbation" and "task-irrelevant perturbation".

To investigate the problem of how the motor system coordinates to adapt the redundant body system to these perturbations, a new experimental approach is required. This study introduces a novel bimanual stick manipulation task specially designed to address this issue (Fig. 2a). Participants manipulated a virtual stick held with both hands to reach visual targets by the tip of the stick. This is a redundant task because the same tip position can be achieved with different stick angles (Fig. 2b). Importantly, this task allowed us to visually perturb the tip position (i.e., end-effector relevant perturbation) and the stick angle (i.e., end-effector irrelevant perturbation) either separately or simultaneously. Using this approach, we can directly investigate how the motor system changes the redundant body system when adapting to these two types of perturbations.

Bimanual stick manipulation task.

(a) Participants held handles of a manipulandum with both hands and moved the right tip of a virtual stick on the monitor. (b) In this task, there are countless movement patterns to move the tip to the target. Two representative strategies are shown: parallel translation (top) and tilting the stick (bottom). (c) During the baseline phase, participants were instructed to move the tip from a starting point to a target that appeared in one of nine directions. (d) Movement trajectories (top) and velocity profiles (bottom) of the left hand, right hand, and tip averaged across participants. Each color corresponds to the target direction shown in c. The shaded areas represent the SEM across participants. (e) The time- course of the stick-tilt angle shows that participants moved the tip while tilting the stick. The shaded areas indicate the SEM. (f) Trial-to-trial changes in stick-tilt angle at the peak velocity of the tip. The strategy of tilting the stick was observed from the beginning of the experiment (t-test compared with zero, |t(71)| > 6.4, Bonferroni corrected P < 0.001). The shaded areas indicate the SEM.

Results

Movement strategy to manipulate the stick

A total of 72 healthy participants took part in the study (Experiments 1–3). They were asked to manipulate a virtual stick with both hands (length = 40 cm; distance between hands = 15 cm) to move the right tip of the stick towards the targets (Fig. 2a). This task is redundant because the same target can be reached in countless movement patterns. The participants could move both hands in parallel without tilting the stick (Fig. 2b, top). Alternatively, they could perform the task by tilting the stick (Fig. 2b, bottom). During the first part of the task (baseline phase: 360 trials) in all the experiments (Experiments 1–3), the participants moved the tip from a starting point to a target (movement distance = 10 cm) that appeared in one of nine directions (0°: horizontal direction, ±10°, ±20°, ±30°, and ±40°) (Fig. 2c).

The trajectories of the tip and both hands appeared to be almost straight, but both hands were not moved in parallel, except when aiming at the 0° target (Fig. 2d, top). The velocity profiles of the hands and tip were bell-shaped (Fig. 2d, bottom), similar to a normal unimanual reaching movement (Flash & Hogan, 1985; Gordon et al., 1994). The time-course changes in the stick-tilt angle relative to the initial horizontal tilt angle showed that the participants moved the tip while slightly tilting the stick, except when aiming at the 0° target (Fig. 2e). This tendency was observed from the beginning of the experiment (t-test compared with 0°, |t(71)| > 6.4, Bonferroni corrected P < 0.001; Fig. 2f).

Movement strategies to manipulate the stick for individuals are illustrated on the plane expressed by two variables: tip-movement direction relative to a horizontal direction (TMD) and the stick-tilt angle relative to the initial horizontal state (STA) (Fig. 3a). Although each participant could have performed the task differently (Fig. 2b), all participants showed stereotypical movement patterns: The stick-tilt angle was significantly modulated by the tip-movement direction (t-test compared with 0°, |t(71)|5.5, Bonferroni corrected P < 0.01; Fig. 3a, inset). This monotonically increasing curve, which will be referred to as the “baseline TMD–STA relationship”, reflects a constraint between the direction of tip movement and the angle of stick tilt angle. The relationship showed that clockwise (CW) tip movement direction was accompanied by CW stick-tilt, and counterclockwise (CCW) tip-movement direction was accompanied by CCW stick-tilt. This constraint of movement patterns was likely created by minimizing the movement distance of the hands. If both hands are moved in parallel, the total distance between the initial and final positions of the hands should be 20 cm (10 cm for each) regardless of the targets; however, the experimentally observed distance was reduced by tilting the stick (Fig. 3b). The average distance was a near-minimal value (versus parallel translation: Cohen’s d = 1.914, versus theoretical minimum: Cohen’s d = 0.334; Fig. 3c).

Relationship between the tip-movement direction and stick-tilt angle.

(a) A relationship between each individual’s tip-movement direction and stick-tilt angle, when the tip speed peaked is represented as the thin gray lines drawn with colored plots. The averaged relationship is plotted as a bold gray line (baseline TMD–STA relationship) with the SD (crosses) and 95% confidence intervals (ellipses). The inset shows the stick-tilt angles at peak velocity and movement offset (error bars: SD, *: t-test compared with 0°, Bonferroni-corrected P < 0.01). The monotonically increasing curve indicates the stereotypical strategy of manipulating the stick (diagrams on the right side).

(b) Sum of the distances between the initial and final positions of both hands (i.e., the sum of the lengths of pink arrows in Fig. 2b). The dotted and dashed gray lines indicate the distances when both hands are moved in parallel and the mathematically derived minimum distances by tilting the stick, respectively. The error bars denote the SD. (c) The distance averaged across target directions was closer to the minimum value (Cohen’s d = 1.914) than the parallel translation distance (Cohen’s d = 0.334). The error bar indicates the SD across participants.

Experiment 1: Motor adaptation pattern to the end-effector relevant perturbation

During the second part of the experiment (adaptation phase: 240 trials), the target appeared only in the 0° direction, and three types of visual perturbations were introduced. Most importantly, this task enabled the introduction of end-effector relevant perturbation (Experiment 1), or end-effector irrelevant perturbation (Experiment 2), or a combination of both types of perturbations (Experiment 3). In Experiment 1A (N = 19), the tip-movement direction was visually rotated in the CCW direction around the starting position (tip perturbation), whereas the stick-tilt angle remained unchanged (Fig. 4a). The amount of tip directional rotation was gradually increased by 1° per a trial up to 30° so that the participants did not notice the presence of perturbation.

Experiment 1: End-effector relevant visual perturbations.

(a) During the adaptation phase of Experiment 1, the tip-movement direction was rotated by 30° around the starting point (tip perturbation). The dotted and solid black arrows indicate the intended and actual tip trajectory, respectively. (b) The changes in the movement pattern are expressed as the state shift on the TMD–STA plane (left: visual space, right: physical space). The dashed red lines denote the task goal (i.e., a target direction). On the visual TMD–STA plane, the perturbation (a blue arrow) shifts the movement pattern. On the physical TMD–STA plane, the task goal is shifted. (c) The participants could correct the tip error without (pattern i, top) or by (pattern ii, bottom) altering the stick-tilt angle. (d) The adaptive strategies are expressed as the shift on the visual TMD–STA plane (left) and the physical TMD–STA plane (right). The visual error caused by the perturbation could be compensated by different adaptation patterns (blue arrows): Representative patterns are expressed as bold arrows (patterns i and ii).

Figure 4b illustrates how this visual perturbation changes the states of stick movement in visual and physical spaces. The movement in the visual space was the one visible to the participants, whereas the movement in the physical space was the actual hand movement, which was invisible to them. On the visual TMD–STA plane (Fig. 4b, left), the visual perturbation to the tip can be expressed as a shift of the state along the TMD axis. The error of the tip-movement direction in the visual space needs to be compensated by changing the movement of both hands. On the physical TMD– STA plane (Fig. 4b, right), this requirement can be expressed as a shift of the task goal in the direction opposite to the visual perturbation.

Owing to the redundant nature of this task, participants could adapt to change the visual tip movement using different strategies. For example, they could change the tip movement without changing the stick-tilt while keeping the strategy of moving both hands in parallel (Fig. 4c, top). Alternatively, they might adapt by changing the stick tilt as if they were aiming at a -30° target (Fig. 4c, bottom). These strategies can be expressed as a state shift towards the new task goal line along the TMD-axis or along the baseline TMD–STA relationship in the physical TMD–STA plane (Fig. 4d).

The experimental results demonstrated that the participants adapted to the tip perturbation by tilting the stick. On the visual TMD–STA plane (Fig. 5a), after the adaptation, the tip-movement direction remained at the baseline level (paired t-test, t(18) = -1.320, P = 0.203), while the stick was tilted significantly more in the CW direction than when aiming at the 0° target in the baseline phase (paired t-test, t(18) = 10.003, P < 0.01). Thus, the tip perturbation was compensated by adaptation, but the adaptation produced an abnormal visual stick-tilt. This result is likely to support the idea of the minimal intervention principle because participants ignored this visual stick-tilt which did not directly influence the task.

Experiment 1: Adaptation patterns to gradually imposed end-effector relevant perturbations.

(a) In Experiment1A, the adaptation trajectory (a thin blue line) on the visual TMD–STA plane gradually deviated from a baseline pattern (a gray circle) mainly along the STA-axis. The tip-movement direction remained at the baseline level (paired t-test, t(18) = -1.320, P = 0.203). In contrast, the stick-tilt angle after the adaptation (a blue circle) significantly differed from that in the baseline pattern when aiming at the 0° target (a gray circle) (paired t-tests, t(18) = 10.003, P < 0.001), indicating that the participants saw the abnormal visual stick-tilt. The plots indicate mean ± SD across participants. (b) On the physical TMD–STA plane, the adaptation trajectory followed the baseline TMD–STA relationship. Although the difference was small but significant for the tip-movement direction compared with the baseline pattern when aiming at the -30° target (paired t-test, t(18) = -10.053, P < 0.01), there was no significant difference in the stick-tilt angle (paired t-test, t(18) = -1.081, P = 0.294). (c) The area indicated by the squared outline in b is enlarged. The adaptive process can be divided into nine trial segments (trials 1–5, 6–10, 11–15, 16–20, 21–25, 26–30, 30–120, 121–210, and 211–240). (d) The motor system implicitly achieved adaptation to the tip perturbation as if it had voluntarily aimed at the -30° target. (e) Participants who showed a larger stick-tilt when aiming at the -30° target during the baseline phase also exhibited a larger stick-tilt in the adapted pattern (R = 0.599, P < 0.001).

On the physical TMD–STA plane, adaptation appeared to progress along the baseline TMD–STA relationship throughout the adaptation phase (Figs. 5b and 5c). The implicitly adapted tip-movement direction and stick-tilt angle were similar to those when voluntarily aiming at the -30° target in the baseline phase. Although the difference was small but significant for the tip-movement direction (paired t-test, t(18) = -10.053, P < 0.01), there was no significant difference in the stick-tilt angle (paired t-test, t(18) = -1.081, P = 0.294). Thus, the motor system implicitly achieved adaptation to the tip perturbation as if it had voluntarily aimed at the -30° target (i.e., pattern ii in Figs. 4c and 4d) (Fig. 5d). This adaptation characteristic was also supported by the observation that the participants who tilted their stick more when aiming at the -30° target in the baseline phase exhibited a greater stick- tilt in the adapted movement patterns (R = 0.601, P = 0.007; Fig. 5e). In Experiment 1B (N = 13), we imposed the 30° tip perturbation abruptly to allow participants to use explicit strategy, and obtained nearly identical results (Fig. S1).

As described above, the adapted tip-movement direction on the physical TMD–STA plane was significantly different from the tip-movement direction when aiming at the -30° target in the baseline phase (Fig. 5b). In contrast, such significant difference was not observed for Experiment 1B (Fig. S1). Why was such a difference only in Experiment 1A? We speculated that this difference was caused by the implicit adaptation to correct the abnormal visual stick-tilt angle. As shown in Fig. 5a, after the adaptation, the stick was visually tilted in the CW direction. If the physical movement pattern is constrained by the baseline TMD–STA relationship, the CCW correction of the stick-tilt angle should be accompanied by the CCW change in tip- movement, which would interfere with the adaptation to the tip perturbation.

Experiment 2: Motor adaptation pattern to end-effector irrelevant perturbation

Thus, it is still possible that the motor system does not necessarily ignore, but corrects for, the stick-tilt error that does not directly influence the tip position. Experiment 2 (N = 20) was designed to directly test this possibility by imposing a stick-tilt perturbation around the right tip (i.e., end-effector irrelevant perturbation) (Fig. 6a). As the tip moved from the starting point, the stick started to tilt around the tip up to ±6° when the tip reached the target (the tilt amount was proportional to the tip’s distance from the starting point). The ±6° tilt perturbation was chosen according to the stick tilt at the movement offset when reaching the ±30° target for the baseline phase (see Fig. 3a). Ten participants participated in the CCW stick-tilt perturbation condition (E2CCW group: 6°), and the remaining ten participants took part in the CW stick-tilt perturbation condition (E2CW group: -6°). Because the amount was small, the participants did not notice the presence of perturbation throughout the experiment. Since this perturbation changed only the stick-tilt angle, it is expressed as a state shift along the STA-axis on the visual TMD–STA plane (Fig. 6b). According to the minimal intervention principle, the motor system does not need to change anything because the stick-tilt perturbation does not affect the tip position (i.e., the task goal) (Figs. 6c and 6d, pattern i). However, if the motor system is actively compensating for the stick-tilt perturbation, it should change the stick-tilt angle (Figs. 6c and 6d, pattern ii).

Experiment 2: End-effector irrelevant visual perturbations.

(a) During the adaptation phase of Experiment 2, the stick-tilt angle was rotated around the tip. The perturbation of ±6° was imposed from the first trial. Participants received either the counterclockwise perturbation (green, E2CCW group) or the clockwise perturbation (orange, E2CW group). (b) On the visual TMD–STA plane, the perturbation is expressed as a vertical shift of the state along the task goal (dashed red line). On the physical TMD–STA plane, the perturbation does not influence the task goal. Note that, in both planes, the states are on the task goal lines. The formats are the same as in Fig. 4b. (c) Participants may not change the movement pattern because the perturbation does not affect the tip position (pattern i, top). Alternatively, the perturbation can lead to a stick-tilt correction (pattern ii, bottom). (d) These strategies are expressed as a shift on the TMD–STA plane. Note that pattern i means that there is no change in the movement pattern.

On the visual TMD–STA plane, the stick-tilt angles remained different from those in the baseline phase (Fig. 7a) (two-way (direction of stick-tilt perturbation (i.e., groups), trial phase (first 30 trials or last 30 trials)) repeated-measures ANOVA, a significant main effect of groups: F(1,18) = 54.457, P < 0.001; no effect of trial phase: F(1,18) = 2.554, P = 0.127; no interaction: F(1,18) = 1.613, P = 0.220). Thus, the adaptive changes did not fully compensate for the perturbation-induced stick-tilt errors. However, the data on the physical TMD–STA plane clearly demonstrated that participants implicitly changed the stick-tilt angle in the direction opposite to that of the perturbation from the early adaptation phase (Fig. 7b): Without the correction, the data of the E2CCW and E2CW groups should have largely overlapped. However, with respect to stick-tilt angle, two-way repeated-measures ANOVAs revealed the main effect of stick-tilt perturbation direction (F(1, 18) = 12.483, P < 0.01) but not the main effect of trial phase (F(1, 18) = 3.331, P = 0.085). The interaction between these factors was not significant (F(1, 18) = 1.362, P = 0.258).

Experiment 2: Adaptation patterns to end-effector irrelevant visual perturbations.

(a) Movement patterns are shown by green (E2CCW group) and orange (E2CW group) plots on the visual TMD–STA plane. The open and filled plots represent the movement patterns in the first and last 30 trials, respectively, during the adaptation phase (mean ± SD, across participants). (b) On the physical TMD–STA plane, the separated movement patterns between both groups showed that participants corrected the end-effector irrelevant stick-tilt errors. (c) The enlarged plot of the squared area in b shows that end-effector relevant tip-movement directional errors accompanied the adaptation. These effects persisted until the end of the adaptation phase (two-way repeated-measures ANOVAs). (d) Schematics of how the stick-tilt perturbation altered the movement patterns. (e, f) How participants tilted the stick during the baseline phase was negatively correlated with the stick-tilt angle and the tip-movement direction. To control for the individual variability, the baseline stick- tilt angle or tip-movement direction when aiming at the 0° target was subtracted from each value.

Unexpectedly, this stick-tilt correction was accompanied by undesirable tip- movement directional errors (Fig. 7c). With respect to the tip-movement direction, two-way repeated-measures ANOVAs revealed the main effect of the direction of stick-tilt perturbation (F(1, 18) = 5.104, P = 0.037) but not the main effect of trial phase (F(1, 18) = 1.288, P = 0.271) (the interaction between these factors was not significant: F(1, 18) = 0.804, P = 0.382). Notably, these seemingly irrational adaptation patterns were consistent with the patterns predicted from the baseline physical TMD– STA relationship: CCW (CW) stick-tilt adaptation to the CW (CCW) stick-tilt perturbation slightly shifted the tip-movement direction in the CCW (CW) direction (Fig. 7d).

We also examined how the amount of adaptation depended on the amount of stick-tilt when voluntarily aiming at the ±30° target in the baseline phase. The 6° stick- tilt perturbation should be insignificant for participants who tilt the stick at a larger angle; thus, this perturbation tends to be ignored. Conversely, for participants who did not tilt the stick, a 6° stick-tilt perturbation should be significant and must be corrected. Figures 7e and 7f show how the amount of stick-tilt correction (Fig. 7e) and tip- movement directional change (Fig. 7f) depended on the amount of stick-tilt angle when aim at the ±30° target in the baseline phase. As predicted, we observed a negative correlation (stick-tilt correction: R = -0.572, P < 0.01; tip-movement directional change: R = -0.440, P = 0.052), although the tip-movement direction change was not significant.

Taken together, the results of Experiment 1 and Experiment 2 suggest that the implicit adaptation in the redundant stick-manipulating task is achieved by the following two mechanisms: First, the motor system attempts to correct for any perturbation in visual feedback, regardless of whether it is end-effector relevant or end-effector irrelevant. Second, the physical pattern of movement correction is constrained by the baseline TMD–STA relationship.

Experiment 3: Interaction between end-effector relevant and end-effector irrelevant perturbations

Experiment 3 (N = 20) was designed to further investigate the interaction between adaptations to end-effector relevant and end-effector irrelevant perturbations by applying both simultaneously (Fig. 8a). We hypothesized that adaptation to the tip perturbation would be influenced by the direction of the simultaneously applied stick- tilt perturbation. Specifically, the CCW tip perturbation was combined with the CCW stick-tilt perturbation (E3CCW group: N = 10) or with the CW stick-tilt perturbation (E3CW group: N = 10). These visual perturbation patterns can be expressed as state shifts in the visual TMD–STA plane (green and orange arrows for E3CCW and E3CW groups, respectively, in Fig. 8b). On the physical TMD–STA plane, the correction pattern required for the E3CCW group (Fig. 8c, top) is aligned with the baseline TMD–STA relationship (green arrow in Fig. 8d), from which we can predict that adaptation should proceed smoothly. In contrast, the correction pattern required for the E3CW group (Fig. 8c, bottom) is orthogonal to the baseline TMD–STA relationship (orange arrow in Fig. 8d), which may prevent smooth adaptation. We tested these predictions by simultaneously imposing the tip perturbation which was the same as the perturbation used in Experiment 1A and the stick-tilt perturbation which was the same as the perturbation used in Experiment 2. Again, the participants were unaware of the perturbations throughout the experiment.

Experiment 3: Interaction between end-effector relevant and end- effector irrelevant visual perturbations.

(a) During the adaptation phase of Experiment 3, both the CCW tip perturbation and the CCW or CW stick-tilt perturbation were introduced. The participants were assigned to the CCW stick-tilt rotation condition (green, E3CCW group) or the CW stick-tilt rotation condition (orange, E3CW group). (b) These visual perturbations can be expressed as state shifts on the visual TMD–STA plane (green and orange arrows for E3CCW and E3CW groups, respectively). (c) The adaptation pattern to reduce visual errors (top: E3CCW group, bottom: E3CW group). (d) On the physical TMD–STA plane, the correction pattern required for the E3CCW group is aligned with the baseline TMD–STA relationship (green arrow), whereas the correction pattern required for the E3CW group is orthogonal to the baseline TMD–STA relationship (orange arrow).

Contrary to the predictions, the difference in stick-tilt perturbation directions did not affect the final adapted movement patterns: As shown in the visual TMD–STA plane (Fig. 9a), the difference in stick-tilt angle remained unchanged (paired t-test, t(18) = 6.920, P < 0.001). Similarly, in the physical TMD–STA plane (Fig. 9b), there were no significant differences in movement pattern between the two groups (tip- movement direction: paired t-test, t(18) = -0.479, P = 0.638, stick-tilt angle: paired t- test, t(18) = -0.386, P = 0.704). However, a trial-by-trial variability of the movement pattern in the latter adaptation phase (trial# 121–240) showed a slight difference between groups (t-test, tip-movement direction: t(18) = -2.046, P = 0.056, stick-tilt angle, t(18) = -3.235, P = 0.005), suggesting that the adaptation processes were implicitly affected by combination of perturbations (Fig. 9a, inset). In partial agreement with our predictions, two groups showed different adaptation processes: trial- dependent trajectories to the final adapted pattern were clearly separated for both groups (Fig. 9c). Indeed, a repeated-measures multivariate analysis of variance (MANOVA) showed a significant effect of group (Pillai’s Trace F(8, 162) = 0.155, P <0.001) and trial segment (Pillai’s Trace F(8, 162) = 0.719, P < 0.001); however, no significant interaction was observed (Pillai’s Trace F(8, 162) = 0.039, P = 0.982). As a result, the adaptation of the tip movement was more delayed in the E3CW group (trial segment 2: t-test, t(18) = -3.684, Bonferroni-corrected P < 0.05; Fig. 9d).

Experiment 3: Adaptation patterns to end-effector relevant and end- effector irrelevant visual perturbations.

(a) Adaptation patterns on the visual TMD– STA plane are shown (same formats as Fig. 5a). A green (orange) plot represents the adapted movement pattern in the E3CCW group (E3CW group). The tip-movement direction did not significantly differ (t-test, t(18) = -0.473, P = 0.642), while the stick-tilt angle remained a significant difference between groups (t-test, t(18) = 6.920, P < 0.001). A trial-by-trial variability of the movement pattern in the latter adaptation phase (trial# 121–240) showed a difference between groups (t-test, tip-movement direction: t(18) =-2.046, P = 0.056, stick-tilt angle, t(18) = -3.235, P = 0.005) (b) Similarly, the adapted movement patterns on the physical TMD–STA plane did not show a significant difference (t-test, tip-movement direction: t(18) = -0.479, P = 0.638, stick-tilt angle: t(18) = -0.386, P = 0.704). (c) However, the adaptive processes appeared to be different. The movement patterns of the two groups are compared using a repeated- measures MANOVA. A multivariate analysis revealed a significant effect of group (Pillai’s trace F(8, 162) = 0.155, P < 0.001) and trial segment (Pillai’s trace F(8, 162) = 0.719, P < 0.001); no significant interaction can be observed (Pillai’s trace F(8, 162) = 0.039, P = 0.982). The definition of trial segments is the same as that shown in Fig. 5c. (d) The learning curves for the tip-movement. The adaptation in the E3CW group was always delayed but showed a significant difference only during the early phase (trial segment 2: t-test, t(18) = -3.684, Bonferroni-corrected P < 0.05).

Discussion

Our bodies have more degrees of freedom than necessary to achieve a relatively low-dimensional task goal. Understanding how the motor system resolves this redundancy remains a challenging problem (Bernstein et al., 1996). In this study, we investigated the problems of motor control and adaptation specific to kinematically redundant systems. The first problem is how the motor system coordinates the redundant body movements to control the end-effector (Fig. 1f). The second problem is how the motor system coordinates to modify redundant body movements when implicitly adapting to the perturbation imposed on the end-effector (Fig. 1i). The third problem is how the motor system responds to the perturbation to the redundant dimension that does not directly affect the end-effector (Fig. 1k). These two types of perturbations have been conventionally referred to as “task-relevant” and “task- irrelevant” perturbations. In previous studies, task-irrelevant perturbations were created by applying the perturbation to the end-effector (e.g., cursor) only at the midpoint of the movement (Wolpert et al., 1995; Flanagan & Rao, 1995) or by using a wide target (e.g., arc) (Schaefer et al., 2012; Kim et al., 2019). Because the task- irrelevant conditions realized in these studies are completely different from the scope of our study, we specifically used the terms "end-effector relevant perturbation" and "end-effector irrelevant perturbation".

To investigate these issues, we designed a novel stick manipulation reaching task. This redundant motor task allows the investigation of adaptation patterns in the redundant system following the introduction of end-effector relevant, end-effector irrelevant, or both types of perturbations. First, we showed that participants adopted a stereotyped strategy to reach the target with the stick tip, and that the movement pattern (stick tip movement and stick tilt angle) was constrained by a baseline TMD– STA relationship. Second, when only end-effector relevant perturbations were imposed on the tip, the motor system attempted to modify the physical movement patterns along the baseline TMD–STA relationship to achieve the task goal (Experiment 1). Third, even when perturbations were applied only to the end-effector irrelevant dimension (i.e., the stick-tilt angle), the motor system still adapted to the end-effector irrelevant perturbation and this adaptation was accompanied by an end- effector relevant error (Experiment 2). Finally, when both types of perturbations were imposed simultaneously, the adaptation processes were influenced depending on the combination of perturbations (Experiment 3).

Kinematic relationship between end-effector relevant and end-effector irrelevant dimensions

In our stick manipulation task, participants were asked to reach the tip of the stick to the targets. Participants can achieve the reaching task in different ways. For example, they can use a simple strategy by moving both hands in the same direction (Fig. 2b, top) or a more complicated strategy by tilting the stick (Fig. 2b, bottom). Nevertheless, they showed stereotypical patterns of moving the stick tip by tilting the stick from the beginning of the task (Fig. 2f): Reaching the CW (or CCW) targets was accompanied by the CW (or CCW) stick tilt (Fig. 3a). Notably, compared with the case where both hands moved in parallel, the movement distances of both hands to perform the task were significantly reduced to a near-optimal level (Figs. 3b and 3c), which contributed to the reduction of work required (Collins, 1995; O’Sullivan et al., 2009). As a result of this optimization process, tip-movement and stick-tilt were uniquely associated with each other by the baseline TMD–STA relationship (Fig. 3a). Physically, this relationship constrains how the motor system should move the stick. Visually, it tells the motor system how the stick movement should be seen.

The strength of the task to investigate the motor adaptation in redundant system

Our task has several advantages. First, the task has created redundancy in a natural way. Manipulating a stick with both hands is relatively common in everyday life (using a broom, golf or bat swing etc.). This familiarity with the task was reflected in Fig. 2f which shows that all participants started tilting the stick to reach the tip toward the targets from the beginning of the experiment. This characteristic is ideal for investigating how this inherent movement pattern influences motor adaptation. Second, this task allows us to investigate how the perturbation to the tip-movement direction leads to a change in both the tip-movement direction itself and the stick-tilt angle. It has not been investigated how the motor system implicitly achieves adaptation to the end-effector perturbation (i.e., the end-effector relevant perturbation) by changing the movement patterns of the redundant body system. Third, this task can create a novel condition in which the perturbations are imposed on the redundant system while the tip position remains unchanged (i.e., the end- effector irrelevant perturbation). No previous studies have investigated how the motor system implicitly responds to such a perturbation by altering (or not altering) the movement of the redundant body system.

Adaptation to end-effector relevant perturbation

When the end-effector relevant perturbation was implicitly imposed to the tip- movement direction (visual rotation by 30° CCW), the participants adapted to change the tip-movement direction in the opposite CW direction (Experiment 1), as reported in many previous studies using an ordinary reaching movement (Cunningham, 1989; Krakauer et al., 2000). Importantly, our task helped to investigate how such motor adaptations in the end-effector relevant dimension (tip-movement direction) are accompanied by the movement pattern changes in the end-effector irrelevant dimension (stick-tilt angle). Participants implicitly corrected the tip-movement direction by changing the stick tilt as if they were voluntarily aiming at the -30° target. This adaptation can be visualized as a state shift along the baseline TMD–STA relationship in the physical TMD–STA plane (Fig. 5b). The baseline TMD–STA relationship originally contributes to dictate how the motor system voluntarily moves the stick. It is likely that the motor system also uses this relationship as a scaffold to implicitly guide how to correct the tip-movement direction. Hayashi et al. (2016) have demonstrated that the adaptation of the standard reaching movement to a visual perturbation was significantly influenced by the baseline relationship between the movement and target directions (i.e., a visuomotor map). After distorting the visuomotor map so that the movement direction was more (or less) sensitively modulated by the target direction, the movement adaptation was enhanced (or reduced). The present finding that the baseline movement pattern is used for the implicit adaptation is consistent with this previous finding.

Notably, this adaptation pattern changed the visual feedback of the stick tilt. When reaching the 0° target in the baseline condition, the stick-tilt angle was almost 0° (Fig. 3a). However, after adaptation, the amount of stick-tilt was equivalent to that when aiming at the -30° target (Fig. 5b). Thus, the visually observed stick-tilt after adaptation was different from that observed in the baseline condition (Fig. 5a). Nevertheless, the adapted physical movement pattern persisted to follow the baseline TMD–STA relationship, suggesting that the motor system mainly paid attention to end-effector relevant tip-movement and ignored end-effector irrelevant stick-tilt angle. This adaptation pattern seems to be consistent with the minimal intervention principle.

However, a slight difference in the adaptation patterns between Experiment 1A and Experiment 1B implies a violation of the minimal intervention principle. In Experiment 1A, in which a 30° visual rotation was implicitly introduced by gradually increasing its amplitude, the final adapted tip-movement direction was slightly but significantly different from that when voluntarily aiming at the adapted direction (i.e., -30° target) (Fig. 5b). In contrast, no such residual difference was not observed in Experiment 1B, where participants could use an explicit aiming strategy (Fig. S1b). Visuomotor adaptation is often incomplete even after sufficient training, and residual errors are observed (Krakauer et al., 2000; Ethier et al., 2008; Malone et al., 2011; Vaswani et al., 2015). Although the presence of residual errors has been attributed to the imbalance between learning and forgetting, a recent study demonstrated that only the implicit learning system is responsible for producing residual errors by modulating error sensitivity with the consistency of errors in previous trials (Albert et al., 2021). We speculate that an alternative mechanism is involved in the generation of a residual difference in Experiment 1A. If the motor system does not ignore the abnormal stick-tilt but attempts to correct it, and this correction pattern is physically constrained by the baseline TMD–STA relationship, the stick-tilt correction should accompany the de-adaptation of the tip-movement direction and lead to the generation of a residual error. As discussed below, the results of Experiment 2 and Experiment 3 support this idea.

Adaptation to end-effector irrelevant perturbation

Our task can create a novel end-effector irrelevant stick-tilt perturbation and allows us to study how the motor system responds to it. According to the minimal intervention principle, the motor system should ignore stick-tilt perturbations because they do not influence task goal achievement (i.e., so called “task-irrelevant” perturbation). However, inconsistent with this idea, Experiment 2 clearly demonstrated that the motor system responded to the stick-tilt perturbations (Figs. 7b and 7c). This result indicates that the motor system predicts the visual feedback of an end-effector irrelevant stick-tilt angle when performing the task and tries to eliminate the dissociation between the observed and predicted stick-tilt angle (i.e., sensory prediction error). This idea was corroborated by the fact that participants who demonstrated a shallower TMD–STA relationship during the baseline phase showed a greater amount of stick-tilt correction (Fig. 7e). The ±6° stick-tilt perturbation should cause the larger prediction error for these participants.

One might consider that previous studies have already shown that the motor system responds to the task-irrelevant perturbations (Wolpert et al., 1995; Flanagan & Rao, 1995; Diedrichsen et al., 2010; Schaefer et al., 2012; Kim et al., 2019). However, the task-irrelevant conditions implemented in these previous studies are quite different from the end-effector irrelevant condition in our study. These studies created the task-irrelevant condition by combining the perturbation of visual cursors with the manipulation of movement trajectory (Wolpert et al., 1995; Flanagan & Rao, 1995; Morehead et al., 2017) or target/cursor shape (Diedrichsen et al., 2010; Schaefer et al., 2012; Kim et al., 2019). Therefore, these perturbations are not “end- effector irrelevant”. Exceptionally, in a previous study, an end-effector irrelevant perturbation was implemented in a bimanual postural task to investigate postural reflex activity (Omrani et al., 2013). However, to our knowledge, such an end-effector irrelevant perturbation has never been used to investigate the motor adaptation pattern in a redundant system. Furthermore, as discussed in the next section, our task design revealed a novel aspect of motor adaptation to the end-effector irrelevant perturbation.

Adaptation to end-effector irrelevant perturbation influences end-effector relevant performance

Unexpectedly, Experiment 2 shows that adaptation to the end-effector irrelevant stick-tilt perturbation was accompanied by an increase in the end-effector relevant tip-movement direction error (Fig. 7c). This counterintuitive result can be explained by the physical constraint of the adaptation pattern along the baseline TMD–STA relationship (Fig. 3a). According to this constraint, CCW (or CW) stick-tilt adaptation should change the tip-movement direction in the CCW (or CW) direction (Fig. 7d). The influence of stick-tilt perturbation on tip-movement direction error tended to be more pronounced for participants with a smaller stick-tilt angle at baseline, i.e., those with a shallower baseline TMD–STA relationship (Fig. 7f), further supporting this idea. The results of Experiment 3 also support the existence of such an interaction between tip-movement and stick-tilt adaptation. When the tip-movement and stick-tilt perturbations were imposed simultaneously, the adaptation trajectories on the physical TMD–STA plane were significantly influenced by the perturbation combination patterns (Fig. 9c). If the motor system had simply attempted to correct the stick-tilt to reduce the sensory prediction error, this spillover effect on the tip- movement direction would not have been observed.

One remaining question is why the motor system needs to correct for end- effector irrelevant stick-tilt errors in the first place? We speculate that there may be two reasons. First, a possible simple explanation is that the motor system predicts all sensory consequences generated by the action and attempts to correct any differences between the predicted and actual sensory information, even when the task is successfully achieved (Mazzoni & Krakauer, 2006; Schaefer et al., 2012; Morehead et al., 2017; Kim et al., 2019). Second, given that tip-movement direction and stick-tilt angle are tightly coupled (i.e., the presence of baseline TMD–STA relationship), the presence of an end-effector irrelevant stick-tilt visual error implies the presence of an end-effector relevant tip-movement direction error. In other words, the motor system interprets end-effector irrelevant errors as evidence that end- effector relevant errors have been also produced. When sensory information contains noise and/or uncertainty, the reliability of sensory information estimation can be improved by utilizing additional sensory information (Ernst & Banks, 2002). From this viewpoint, utilizing stick-tilt information may be useful for improving the accuracy of the estimation of the tip-movement direction.

Figure 10 illustrates the possible mechanism by which the motor system alters the movement pattern of the redundant system when encountering visual errors. The motor system develops a stereotyped movement pattern through an optimization process that is expressed by a baseline TMD–STA relationship (gray lines in Figs. 10a and 10b). This relationship not only dictates how the motor system physically controls the redundant system, but also informs the resulting visual feedback. When visual perturbations are applied to the tip-movement direction or the stick-tilt angle (bold arrows in Fig. 10a), the motor system tries to eliminate the dissociation of the perturbed state from the original state (thin arrows in Fig. 10a). It even responds to the end-effector irrelevant perturbation (red and blue arrows in Fig. 10a), because the error in the end-effector irrelevant dimension implies the presence of an end-effector relevant error (black dashed arrows in Fig. 10a) under the condition that the stick-tilt angle is tightly coupled with the tip-movement direction (i.e., the baseline TMD–STA relationship). However, these correction patterns are not physically allowed. Since the tip-movement direction and the stick-tilt angle are physically constrained by the baseline TMD–STA relationship, these correction patterns are biased toward the relationship (bold arrows in Fig. 10b).

A hypothesis for how the motor system alters the movement pattern of redundant system.

The motor system develops a stereotypical relationship in redundant space, possibly through an optimization process (a gray line). (a) In the visual space, this relationship plays a crucial role for the motor system in how it should implicitly respond to the various visual perturbations (bold arrows) in the redundant space. The motor system tries to eliminate the dissociation of the perturbed state from the original state (thin arrows). According to the relationship, the motor system interprets end-effector irrelevant errors (red or blue) as evidence that end-effector relevant errors have been also produced (dashed arrows). (b) However, the correction patterns are constrained by the relationship in the physical space. The relationship determined which movement pattern the motor system should perform. Thus, the correction patterns gravitate toward the relationship (bold arrows).

In summary, this study shows that the baseline TMD–STA relationship, originally developed to control the redundant system, also plays a crucial role for the motor system in how it should implicitly respond to the visual errors and how it should correct the physical movement. This idea provides a novel framework for understanding how the motor system leans to control the redundant system when encountering visual errors.

Materials and Methods

Participants

A total of 72 healthy adults (25 females and 47 males, aged 22.1 ± 2.5 years old; mean ± SD) participated in the study. Participants were recruited using an online recruitment system (https://www.jikken-baito.com). All the participants were right-handed according to their laterality score (83.3 ± 13.7; mean ± SD) obtained using the Edinburgh Handedness Inventory (Oldfield, 1971). They provided written informed consent, and the experimental protocol was approved by the ethics committee of The University of Tokyo (#19-225).

Experimental apparatus

The participants performed reaching movements using a virtual stick displayed on a horizontal screen with both handles of a manipulandum (KINARM End-Point Lab; Kinarm, Kingston, Canada) (Scott, 1999) (Fig. 2a). This screen prevented the participants from directly seeing their arms and handles, and displayed a start position, a target (1.4 cm in diameter), and a virtual stick (length: 40 cm, width: 0.5 cm). A white circle (1 cm diameter) was displayed on the right tip of the stick. The participants held the left tip of the stick with their left hand (left handle) and the stick 15 cm away from the left hand with their right hand (right handle). The distance between the handles was fixed using a strong elastic force produced by the manipulandum (spring constant = 2000 N/m). The positions and velocities of the hands were analogue/digitally converted at 1.129 kHz and then recorded at 1 kHz for offline analysis.

Baseline phase

All the experiments (Experiments 1–3) comprised a baseline phase (360 trials) followed by an adaptation phase (240 trials). The baseline phase, which was common to all experiments, was performed to investigate the basic strategy of how the participants moved the stick to reach the targets with the right tip. Before each reaching movement, the participants were asked to move the stick to the home position with the help of a manipulandum (Fig. 2c). After maintaining the home position for 0.5−1 s, a gray target was pseudo-randomly chosen from nine possible targets (0°: horizontal direction, ±10°, ±20°, ±30°, ±40°) appearing 10 cm away from the initial right tip position of the stick. After waiting for another 0.5−1 s, the color of the target changed to pink, indicating that they should immediately start moving the tip to the target. The participants were instructed to maintain the peak velocities of the tip as constant as possible across the trials. When the tip reached the target, a feedback message, “fast” or “slow,” was presented on the monitor if the movement speed was higher or lower than the range of 300−450 mm/s, respectively. After maintaining the tip at the target for 0.5 s, the visual feedback of the stick disappeared. Then, the handles automatically returned to their home position, and the next trial began. During the subsequent 240 trials in the adaptation phase, the target appeared only in the 0° direction, and distinct visual perturbations were introduced in Experiments 1–3 as described below. The home positions, size of the visual feedback and the target, inter-trial interval, and criteria for the tip velocity feedback for the adaptation phase were the same as those in the baseline phase.

Experiment 1

Experiment 1A (N = 19) and Experiment 1B (N = 13) were performed to investigate how participants moved the stick when adapting to end-effector relevant perturbations (tip perturbations). In Experiment 1A, the tip-movement direction on the monitor was gradually rotated CCW direction around the starting position by 1° per trial up to 30° (Fig. 4a). The stick-tilt angle remained unchanged (i.e., the stick-tilt angle was identical to the angle of the line connecting both hands). In Experiment 1A, participants did not perceive the presence of rotation and the adaptation was considered to be dominated by the implicit process. On the other hand, in Experiment 1B, the 30° visual rotation was abruptly applied to the tip-movement direction, which allowed them to use an explicit strategy.

Experiment 2

Experiment 2 (N = 20) examined whether the motor system was indifferent to end-effector irrelevant perturbations (stick-tilt perturbations). The stick-tilt angle on the monitor was rotated around the right tip in the CW direction (E2CW group, N = 10) or in the CCW direction (E2CCW group, N = 10). Notably, these perturbations did not affect the tip-movement direction. During each reaching trial, the stick-tilt angle increased from 0° to 6° in proportion to the distance of the tip position from the starting point (0.6°/cm) (Fig. 6a). Participants did not notice any perturbations during the experiment.

Experiment 3

In Experiment 3 (N = 20), to examine how end-effector irrelevant stick-tilt perturbations influenced adaptation to end-effector relevant tip perturbations, both tip- movement and stick-tilt perturbations were simultaneously imposed (Fig. 8a). Ten participants participated in a task in which CCW tip and CCW stick-tilt rotations were imposed (E3CCW group). The remaining ten participants participated in a task in which CCW tip and CW stick-tilt rotations were imposed (E3CW group). The procedure for applying tip and stick-tilt perturbations was the same as in Experiment 1A and Experiment 2, respectively. The participants did not notice perturbations during the experiment.

Data analysis

Data analyses were performed using MATLAB R2023a (MathWorks, Natick, Massachusetts, USA). The behavioral data were low-pass filtered (10 Hz, fourth- order Butterworth filter). The tip-movement direction was calculated as the movement direction of the tip relative to the horizontal direction (Fig. 3a). The stick-tilt angle was defined as the tilting angle of the stick from the horizontal line (Fig. 3a). Positive values for both parameters were defined as the degree to the CCW direction. Unless otherwise noted, the tip-movement direction and the stick-tilt angle at the tip’s peak velocity were used to evaluate the feedforward motor commands.

Minimum distance of hand movement

The total movement distance of both hands in Figs. 3b and 3c was estimated by summing the Euclidean distances between the initial and final positions of the hands. The distance is represented as a function of the stick-tilt angle at the movement offset (𝜃). The first term on the right side represents the distance of the left-hand movement, whereas the second term represents the distance of the right-hand movement. 𝜙, d, L, and R represent a target direction (0°, ±10°, ±20°, ±30°, and ±40°), a target distance (10 cm), a distance between the left hand and the tip (40 cm), and a distance between the right hand and the tip (25 cm), respectively. The minimum distance of hand movement was derived by solving the following equation:

Statistical analysis

To statistically evaluate the behavioral data, we conducted t-tests, correlation analyses, two-way repeated-measures ANOVA, and MANOVA using MATLAB, the software package JASP (http://jasp-stats.org), and R studio.

Data Availability

Data collected in Experiments 1–3 and the code are uploaded in zenodo and are available at https://zenodo.org/record/8033232.

eLife Assessment

This study presents a valuable finding on how the sensorimotor control system deals with redundancy within our body, based on a novel bimanual task. The evidence supporting the authors' claims is convincing, as demonstrated over four different experiment. The work will be of interest to researchers from the motor control community and related fields, and further investigation into the interpretation of the findings could increase the generalisation of the study to a broader audience.

Reviewer #1 (Public Review):

Summary/Strengths:

This manuscript describes a stimulating contribution to the field of human motor control. The complexity of control and learning is studied with a new task offering a myriad of possible coordination patterns. Findings are original and exemplify how baseline relationships determine learning.

Weaknesses:

A new task is presented: it is a thoughtful one, but because it is a new one, the manuscript section is filled with relatively new terms and acronyms that are not necessarily easy to rapidly understand.

First, some more thoughts may be devoted to the take-home message. In the title, I am not sure manipulating a stick with both hands is a key piece of information. Also, the authors appear to insist on the term 'implicit', and I wonder if it is a big deal in this manuscript and if all the necessary evidence appears in this study that control and adaptation are exclusively implicit. As there is no clear comparison between gradual and abrupt sessions, the authors may consider removing at least from the title and abstract the words 'implicit' and 'implicitly'. Most importantly, the authors may consider modifying the last sentence of the abstract to clearly provide the most substantial theoretical advance from this study.

It seems that a substantial finding is the 'constraint' imposed by baseline control laws on sensorimotor adaptation. This seems to echo and extend previous work of Wu, Smith et al. (Nat Neurosci, 2014): their findings, which were not necessarily always replicated, suggested that the more participants were variable in baseline, the better they adapted to a systematic perturbation. The authors may study whether residual errors are smaller or adaptation is faster for individuals with larger motor variability in baseline. Unfortunately, the authors do not present the classic time course of sensorimotor adaptation in any experiment. The adaptation is not described as typically done: the authors should thus show the changes in tip movement direction and stick-tilt angle across trials, and highlight any significant difference between baseline, early adaptation, and late adaptation, for instance. I also wonder why the authors did not include a few no-perturbation trials after the exposure phase to study after-effects in the study design: it looks like a missed opportunity here. Overall, I think that showing the time course of adaptation is necessary for the present study to provide a more comprehensive understanding of that new task, and to re-explore the role of motor variability during baseline for sensorimotor adaptation.

The distance between hands was fixed at 15 cm with the Kinarm instead of a mechanical constraint. I wonder how much this distance varied and more importantly whether from that analysis or a force analysis, the authors could determine whether one hand led the other one in the adaptation.

I understand the distinction between task- and end-effector irrelevant perturbation, and at the same time results show that the nervous system reacts to both types of perturbation, indicating that they both seem relevant or important. In line 32, the errors mentioned at the end of the sentence suggest that adaptation is in fact maladaptive. I think the authors may extend the Discussion on why adaptation was found in the experiments with end-effector irrelevant and especially how an internal (forward) model or a pair of internal (forward) models may be used to predict both the visual and the somatosensory consequences of the motor commands.

Reviewer #2 (Public Review):

Summary:

The authors have developed a novel bimanual task that allows them to study how the sensorimotor control system deals with redundancy within our body. Specifically, the two hands control two robot handles that control the position and orientation of a virtual stick, where the end of the stick is moved into a target. This task has infinite solutions to any movement, where the two hands influence both tip-movement direction and stick-tilt angle. When moving to different targets in the baseline phase, participants change the tilt angle of the stick in a specific pattern that produces close to the minimum movement of the two hands to produce the task. In a series of experiments, the authors then apply perturbations to the stick angle and stick movement direction to examine how either tip-movement (task-relevant) or stick-angle (task-irrelevant) perturbations affect adaptation. Both types of perturbations affect adaptation, but this adaptation follows the baseline pattern of tip-movement and stick angle relation such that even task-irrelevant perturbations drive adaptation in a manner that results in task-relevant errors. Overall, the authors suggest that these baseline relations affect how we adapt to changes in our tasks. This work provides an important demonstration that underlying solutions/relations can affect the manner in which we adapt. I think one major contribution of this work will also be the task itself, which provides a very fruitful and important framework for studying more complex motor control tasks.

Strengths:

Overall, I find this a very interesting and well-written paper. Beyond providing a new motor task that could be influential in the field, I think it also contributes to studying a very important question - how we can solve redundancy in the sensorimotor control system, as there are many possible mechanisms or methods that could be used - each of which produces different solutions and might affect the manner in which we adapt.

Weaknesses:

I would like to see further discussion of what the particular chosen solution implies in terms of optimality.

The underlying baseline strategy used by the participants appears to match the path of minimum movement of the two hands. This suggests that participants are simultaneously optimizing accuracy and minimizing some metabolic cost or effort to solve the redundancy problem. However, once the perturbations are applied, participants still use this strategy for driving adaptation. I assume that this means that the solution that participants end up with after adaptation actually produces larger movements of the two hands than required. That is - they no longer fall onto the minimum hand movement strategy - which was used to solve the problem. Can the authors demonstrate that this is either the case or not clearly? These two possibilities produce very different implications in terms of the results.

If my interpretation is correct, such a result (using a previously found solution that no longer is optimal) reminds me of the work of Selinger et al., 2015 (Current Biology), where participants continue to walk at a non-optimal speed after perturbations unless they get trained on multiple conditions to learn the new landscape of solutions. Perhaps the authors could discuss their work within this kind of interpretation. Do the authors predict that this relation would change with extensive practice either within the current conditions or with further exploration of the new task landscape? For example, if more than one target was used in the adaptation phase of the experiment?

On the other hand, if the adaptation follows the solution of minimum hand movement and therefore potentially effort, this provides a completely different interpretation.

Overall, I would find the results even more compelling if the same perturbations applied to movements to all of the targets and produced similar adaptation profiles. The question is to what degree the results derive from only providing a small subset of the environment to explore.

Reviewer #3 (Public Review):

Summary:

This study explored how the motor system adapts to new environments by modifying redundant body movements. Using a novel bimanual stick manipulation task, participants manipulated a virtual stick to reach targets, focusing on how tip-movement direction perturbations affected both tip movement and stick-tilt adaptation. The findings indicated a consistent strategy among participants who flexibly adjusted the tilt angle of the stick in response to errors. The adaptation patterns are influenced by physical space relationships, guiding the motor system's choice of movement patterns. Overall, this study highlights the adaptability of the motor system through changes in redundant body movement patterns.

Strengths:

This paper introduces a novel bimanual stick manipulation task to investigate how the motor system adapts to novel environments by altering the movement patterns of our redundant body.

Weaknesses:

The generalizability of the findings is quite limited. It would have been interesting to see if the same relationships were held for different stick lengths (i.e., the hands positioned at different start locations along the virtual stick) or when reaching targets to the left and right of a start position, not just at varying angles along one side. Alternatively, this study would have benefited from a more thorough investigation of the existing literature on redundant systems instead of primarily focusing on the lack of redundancy in endpoint-reaching tasks. Although the novel task expands the use of endpoint robots in motor control studies, the utility of this task for exploring motor control and learning may be limited.

Author response:

The following is the authors’ response to the original reviews.

We would like to thank all the reviewers for their positive evaluation of our paper, as described in the Strengths section. We are also grateful for their helpful comments and suggestions, which we have addressed below. We believe that the manuscript has been significantly improved as a result of these suggestions. In addition to these changes, we also corrected some inconsistencies (statistical values in the last sentence of a Figure 5 caption) and sentences in the main text (lines 155, 452, 522) (these corrections did not affect the results).

Fig. 5e: R=0.599, P<0.001 -> R=0.601, P=0.007

L150: "the angle of stick tilt angle" -> "the angle of stick tilt"

L437: "no such" -> "such"

L522: "?" -> "."

Reviewer #1 (Public Review):

Summary/Strengths:

This manuscript describes a stimulating contribution to the field of human motor control. The complexity of control and learning is studied with a new task offering a myriad of possible coordination patterns. Findings are original and exemplify how baseline relationships determine learning.

Weaknesses:

A new task is presented: it is a thoughtful one, but because it is a new one, the manuscript section is filled with relatively new terms and acronyms that are not necessarily easy to rapidly understand.

First, some more thoughts may be devoted to the take-home message. In the title, I am not sure manipulating a stick with both hands is a key piece of information. Also, the authors appear to insist on the term ‘implicit’, and I wonder if it is a big deal in this manuscript and if all the necessary evidence appears in this study that control and adaptation are exclusively implicit. As there is no clear comparison between gradual and abrupt sessions, the authors may consider removing at least from the title and abstract the words ‘implicit’ and ‘implicitly’. Most importantly, the authors may consider modifying the last sentence of the abstract to clearly provide the most substantial theoretical advance from this study.

Thank you for your positive comment on our paper. We agree with the reviewer that our paper used a lot of acronyms that might confuse the readers. As we have addressed below (in the rebuttal to the Results section), we have reduced the number of acronyms.

Regarding the comment on the use of the word “implicit” in the title and the abstract, we believe that its use in this paper is very important and indispensable. One of our main findings was that the pattern of adaptation between the tip-movement direction and the stick-tilt angle largely followed that in the baseline condition when aiming at different target directions. This adaptation was largely implicit because participants were not aware of the presence of the perturbation as the amount of perturbation was gradually increased. This implicitness suggests that the adaptation pattern of how the movement should be corrected is embedded in the motor learning system. On the other hand, if this adaptation pattern was achieved on the basis of the explicit strategy of changing the direction of the tip-movement, the adaptation pattern that follows the baseline pattern is not at all surprising. For these reasons, we will continue to use the word "implicit".

It seems that a substantial finding is the ‘constraint’ imposed by baseline control laws on sensorimotor adaptation. This seems to echo and extend previous work of Wu, Smith et al. (Nat Neurosci, 2014): their findings, which were not necessarily always replicated, suggested that the more participants were variable in baseline, the better they adapted to a systematic perturbation. The authors may study whether residual errors are smaller or adaptation is faster for individuals with larger motor variability in baseline. Unfortunately, the authors do not present the classic time course of sensorimotor adaptation in any experiment. The adaptation is not described as typically done: the authors should thus show the changes in tip movement direction and stick-tilt angle across trials, and highlight any significant difference between baseline, early adaptation, and late adaptation, for instance. I also wonder why the authors did not include a few noperturbation trials after the exposure phase to study after-effects in the study design: it looks like a missed opportunity here. Overall, I think that showing the time course of adaptation is necessary for the present study to provide a more comprehensive understanding of that new task, and to re-explore the role of motor variability during baseline for sensorimotor adaptation.

We appreciate the reviewer for raising these important issues.

Regarding the learning curve, because the amount of perturbation was gradually increased except for Exp.1B, we were not able to obtain typical learning curves (i.e., the curve showing errors decaying exponentially with trials). However, it may still be useful to show how the movement changed with trials during adaptation. Therefore, following the reviewer's suggestion, we have added the figures of the time course of adaptation in the supplementary data (Figures S1, S2, S4, and S5).

There are two reasons why our experiments did not include aftereffect quantification trials (i.e., probe trials). First, in the case of adaptation to a visual perturbation (e.g., visual rotation), probe trials are not necessary because the degree of adaptation can be easily quantified by the amount of compensation in the perturbation trials (however, in the case of dynamic perturbations such as force fields, the use of probe trials is necessary). Second, the inclusion of probe trials allows participants to be aware of the presence of the perturbation, which we would like to avoid.

We also appreciate the interesting additional questions regarding the relevance of our work to the relationship between baseline motor variability and adaptation performance. As this topic, although interesting, is outside the scope of this paper, we concluded that we would not address it in the manuscript. In fact, the experiments were not ideal for quantifying motor variability in the baseline phase because participants had to aim at different targets, which could change the characteristics of motor variability. In addition, we gradually increased the size of the perturbation except for Exp.1B (see Author response image 1, upper panel), which could make it difficult to assess the speed of adaptation. Nevertheless, we think it is worth mentioning this point in this rebuttal. Specifically, we examined the correlation between baseline motor variability when aiming the 0 deg target (tip-movement direction or stick-tilt angle) and adaptation speed in Exp 1A and Exp 1B (Author response image 1 and Author response image 2). To assess adaptation speed in Exp.1A, we quantified the slope of the tip-movement direction to a gradually increasing perturbation (Author response image 1, upper panel). The adaptation speed in Exp.1B was obtained by fitting the exponential function to the data (Author response image 2, upper panel). Although the statistical results were not completely consistent, we found that the participants with greater the motor variability at baseline tended to show faster adaptation, as shown in a previous study (Wu et al., Nat Neurosci, 2014).

Correlation between the baseline variability and learning speed (Experiment 1A).

In Exp 1A, the rotation of the tip-movement direction was gradually increased by 1 degree per trial up to 30 degrees. The learning speed was quantified by calculating how quickly the direction of movement followed the perturbation (upper panel). The lower left panel shows the variability of the tip-movement direction versus learning speed, while the lower right panel shows the variability of the stick-tilt angle versus learning speed. Baseline variability was calculated as a standard deviation across trials (trials in which a target appeared in a 0-degree direction).

Correlation between the baseline variability and learning speed (Experiment 1B).

In Exp 1B, the rotation of the tip-movement direction was abruptly applied from the first trial (30 degrees). The learning speed was calculated as a time constant obtained by exponential curve fitting. The lower left panel shows the variability of the tip-movement direction versus learning speed, while the lower right panel shows the variability of the stick-tilt angle versus learning speed. Baseline variability was calculated as a standard deviation across trials (trials in which a target appeared in a 0-degree direction).

The distance between hands was fixed at 15 cm with the Kinarm instead of a mechanical constraint. I wonder how much this distance varied and more importantly whether from that analysis or a force analysis, the authors could determine whether one hand led the other one in the adaptation.

Thank you very much for this important comment. Since the distance between the two hands was maintained by the stiff virtual spring (2000 N/m), it was kept almost constant throughout the experiments as shown in Author response image 3 (the averaged distance during a movement). The distance was also maintained during reaching movements (Author response image 4).

We also thank the reviewer for the suggestion regarding the force analysis. As shown in Author response image 5, we did not find a role for a specific hand for motor adaptation from the handle force data. Specifically, Author response image 5 shows the force applied to each handle along and orthogonal to the stick. If one hand led the other in adaptation, we should have observed a phase shift as adaptation progressed. However, no such hand specific phase shift was observed. It should be noted, however, that it was theoretically difficult to know from the force sensors which hand produced the force first, because the force exerted by the right handle was transmitted to the left handle and vice versa due to the connection by the stiff spring.

The distance between hands during the task.

We show the average distance between hands for each trial. The shaded area indicates the standard deviation across participants.

Time course changes in the distance between hands during the movement.

The color means the trial epoch shown in the right legend.

The force profile during the movement (Exp 1A).

We decomposed the force of each handle into the component along (upper panels) and orthogonal to the stick (lower panels). Changes in the force profiles in the adaptation phase are shown (left: left hand force, right: right hand force). The colors (magenta to cyan) mean trial epoch shown in the right legend.

I understand the distinction between task- and end-effector irrelevant perturbation, and at the same time results show that the nervous system reacts to both types of perturbation, indicating that they both seem relevant or important. In line 32, the errors mentioned at the end of the sentence suggest that adaptation is in fact maladaptive. I think the authors may extend the Discussion on why adaptation was found in the experiments with end-effector irrelevant and especially how an internal (forward) model or a pair of internal (forward) models may be used to predict both the visual and the somatosensory consequences of the motor commands.

Thank you very much for your comment. As we already described in the discussion of the original manuscript (Lines 519-538 in the revised manuscript), two potential explanations exist for the motor system’s response to the end-effector irrelevant perturbation (i.e., stick rotation). First, the motor system predicts the sensory information associated with the action and attempts to correct any discrepancies between the prediction and the actual sensory consequences, regardless of whether the error information is end-effector relevant or end-effector irrelevant. Second, given the close coupling between the tip-movement direction and stick-tilt angle, the motor system can estimate the presence of end-effector relevant error (i.e., tip-movement direction) by the presence of end-effector irrelevant error (i.e., stick-tilt angle). This estimation should lead to the change in the tip-movement direction. As the reviewer pointed out, the mismatch between visual and proprioceptive information is another possibility, we have added the description of this point in Discussion (Lines 523-526).

Reviewer #1 (Recommendations For The Authors):

Minor

Line 16: “it remains poorly understood” is quite subjective and I would suggest reformulating this statement.

We have reformulated this statement as “This limitation prevents the study of how….” (Line 16).

Introduction

Line 49: the authors may be more specific than just saying ‘this task’. In particular, they need to clarify that there is no redundancy in studies where the shoulder is fixed and all movement is limited to a plane ... which turns out to truly happen in a limited set of experimental setups (for example: Kinarm exoskeleton, but not endpoint; Kinereach system...).

We have changed this to “such a planar arm-reaching task” (Line 49).

Line 61: large, not infinite because of biomechanical constraints.

We have changed “an infinite” to “a large” (Line 61) and “infinite” to “a large number of” (legend in Fig. 1f).

Lines 67-69: consider clarifying.

We have tried to clarify the sentence (Lines 67-69).

Results

TMD and STA, and TMD-STA plane, are new terms with new acronyms that are not easy to immediately understand. Consider avoiding acronyms.

We have reduced the use of these acronyms as much as possible.

“visual TMD–STA plane” -> “plane representing visual movement patterns” (Lines 179180)

“TMD axis” -> “x-axis” (Line 181, Line 190)

“physical TMD–STA plane” -> “plane representing physical movement patterns” (Lines 182-187)

“physical TMD–STA plane” -> “physical plane” (Line 191, Line 201, Lines 216-217, Line 254, Line 301, Line 315, Line 422, Line 511, and captions of Figures 4-9, S3)

“visual TMD–STA plane” -> “visual plane” (Line 193, Line 241, Line 248, Line 300, Lines

313-314, and captions of Figures 4-9, S3)

“STA axis” -> “y-axis” (Line 241)

Line 169: please clarify the mismatch(es) that are created when the tip-movement direction is visually rotated in the CCW direction around the starting position (tip perturbation), whereas the stick-tilt angle remains unchanged.

Thank you for your pointing this out. We have clarified that the stick-tilt angle remains identical to the tilt of both hands (Lines 171-172).

Discussion

I understand the physical constraint imposed between the 2 hands with the robotic device, but I am not sure I understand the physical constraint imposed by the TMD-STA relationship.

The phrase “physical constraint” meant the constraint of the movement on the physical space. However, as the reviewer pointed out, this phrase could confuse the constraint between the two hands. Therefore, we have avoided using the phrase “physical constraint” throughout the manuscript.

Some work looking at 3-D movements should be used for Discussion (e.g. Lacquaniti & Soechting 1982; work by d’Avella A or Jarrasse N).

Thank you for sharing this important information. We have cited these studies in Discussion (Lines 380-382).

Reviewer #2 (Public Review):

Summary:

The authors have developed a novel bimanual task that allows them to study how the sensorimotor control system deals with redundancy within our body. Specifically, the two hands control two robot handles that control the position and orientation of a virtual stick, where the end of the stick is moved into a target. This task has infinite solutions to any movement, where the two hands influence both tip-movement direction and stick-tilt angle. When moving to different targets in the baseline phase, participants change the tilt angle of the stick in a specific pattern that produces close to the minimum movement of the two hands to produce the task. In a series of experiments, the authors then apply perturbations to the stick angle and stick movement direction to examine how either tipmovement (task-relevant) or stick-angle (task-irrelevant) perturbations affect adaptation. Both types of perturbations affect adaptation, but this adaptation follows the baseline pattern of tip-movement and stick angle relation such that even task-irrelevant perturbations drive adaptation in a manner that results in task-relevant errors. Overall, the authors suggest that these baseline relations affect how we adapt to changes in our tasks. This work provides an important demonstration that underlying solutions/relations can affect the manner in which we adapt. I think one major contribution of this work will also be the task itself, which provides a very fruitful and important framework for studying more complex motor control tasks.

Strengths:

Overall, I find this a very interesting and well-written paper. Beyond providing a new motor task that could be influential in the field, I think it also contributes to studying a very important question - how we can solve redundancy in the sensorimotor control system, as there are many possible mechanisms or methods that could be used - each of which produces different solutions and might affect the manner in which we adapt.

Weaknesses:

I would like to see further discussion of what the particular chosen solution implies in terms of optimality.

The underlying baseline strategy used by the participants appears to match the path of minimum movement of the two hands. This suggests that participants are simultaneously optimizing accuracy and minimizing some metabolic cost or effort to solve the redundancy problem. However, once the perturbations are applied, participants still use this strategy for driving adaptation. I assume that this means that the solution that participants end up with after adaptation actually produces larger movements of the two hands than required. That is - they no longer fall onto the minimum hand movement strategy - which was used to solve the problem. Can the authors demonstrate that this is either the case or not clearly? These two possibilities produce very different implications in terms of the results.

If my interpretation is correct, such a result (using a previously found solution that no longer is optimal) reminds me of the work of Selinger et al., 2015 (Current Biology), where participants continue to walk at a non-optimal speed after perturbations unless they get trained on multiple conditions to learn the new landscape of solutions. Perhaps the authors could discuss their work within this kind of interpretation. Do the authors predict that this relation would change with extensive practice either within the current conditions or with further exploration of the new task landscape? For example, if more than one target was used in the adaptation phase of the experiment?

On the other hand, if the adaptation follows the solution of minimum hand movement and therefore potentially effort, this provides a completely different interpretation.

Overall, I would find the results even more compelling if the same perturbations applied to movements to all of the targets and produced similar adaptation profiles. The question is to what degree the results derive from only providing a small subset of the environment to explore.

Thank you very much for pointing out this significant issue. As the reviewer correctly interprets, the physical movement patterns deviated from the baseline relationship as exemplified in Exp.2. However, this deviation is not surprising for the following reason. Under the perturbation that creates the dissociation between the hands and the stick, the motor system cannot simultaneously return both the visual stick motion and physical hands motion to the original motions: When the motor system tries to return the visual stick motion to the original visual motion, then the physical hands motion inevitably deviates from the original physical hands motion, and vice versa.

Our interpretation of this result is that the motor system corrects the movement to reduce the visual dissociation of the visual stick motion from the baseline motion (i.e., sensory prediction error), but this movement correction is biased by the baseline physical hands motion. In other words, the motor system attempts to balance the minimization of sensory prediction error and the minimization of motor cost. Thus, our results do not indicate that the final adaptation pattern is non-optimal, but rather reflect the attempts for optimization.

In the revised manuscript, we have added the description of this interpretation (Lines 515-517).

Reviewer #2 (Recommendations For The Authors):

The authors have suggested that the only study (line 472) that has also examined an end-effector irrelevant perturbation is the bimanual study of Omrani et al., 2013, which only examined reflex activity rather than adaptation. To clarify this issue - exactly what is considered end-effector irrelevant perturbations - I was wondering about the bimanual perturbations in Dimitriou et al., 2012 (J Neurophysiol) and the simultaneous equal perturbations in Franklin et al., 2016 (J Neurosci), as well as other recent papers studying task-irrelevant disturbances which aren’t discussed. I would consider these both to also be end-effector irrelevant perturbations, although again they only used these to study reflex activity and not adaptation as in the current paper. Regardless, further explanation of exactly what is the difference between task-irrelevant and end-effector irrelevant would be useful to clarify the exact difference between the current manuscript and previous work.

Thank you for your helpful comments. We have included as references the study by Dimitriou et al. (Line 490) and Franklin et al. (Lines 486-487), which use an endeffector irrelevant perturbation and the task-irrelevant perturbation condition, respectively. We have also added further explanation of what is the difference between task-irrelevant and end-effector irrelevant (Lines 344-352).

Line 575: I assume that you mean peak movement speed

We have added “peak”. (Line 597).

Reviewer #3 (Public Review):

Summary:

This study explored how the motor system adapts to new environments by modifying redundant body movements. Using a novel bimanual stick manipulation task, participants manipulated a virtual stick to reach targets, focusing on how tip-movement direction perturbations affected both tip movement and stick-tilt adaptation. The findings indicated a consistent strategy among participants who flexibly adjusted the tilt angle of the stick in response to errors. The adaptation patterns are influenced by physical space relationships, guiding the motor system’s choice of movement patterns. Overall, this study highlights the adaptability of the motor system through changes in redundant body movement patterns.

Strengths:

This paper introduces a novel bimanual stick manipulation task to investigate how the motor system adapts to novel environments by altering the movement patterns of our redundant body.

Weaknesses:

The generalizability of the findings is quite limited. It would have been interesting to see if the same relationships were held for different stick lengths (i.e., the hands positioned at different start locations along the virtual stick) or when reaching targets to the left and right of a start position, not just at varying angles along one side. Alternatively, this study would have benefited from a more thorough investigation of the existing literature on redundant systems instead of primarily focusing on the lack of redundancy in endpointreaching tasks. Although the novel task expands the use of endpoint robots in motor control studies, the utility of this task for exploring motor control and learning may be limited.

Thank you very much for the important comment. Given that there are many parameters (e.g., stick length, locations of hands, target position etc), one may wonder how the findings obtained from only one combination can be generalized to other configurations. In the revised manuscript, we have explicitly described this point (Lines 356-359).

Thus, the generalizability needs to be investigated in future studies, but we believe that the main results also apply to other configurations. Regarding the baseline stick movement pattern, the control with tilting the stick was observed regardless of the stick-tip positions (Author response image 6). Regarding the finding that the adapted stick movement patterns follow the baseline movement patterns, we confirmed the same results even when the other targets were used as the target for the adaptation (Author response image 7).

Stick-tip manipulation patterns when the length of the stick varied.

Patterns of adaptation when using the other targets.

In the baseline phase, 40 naïve participants moved a stick tip to a peripheral target (24 directions). They showed a stereotypical relationship between the tip-movement direction and the stick-tilt angle (a bold gray curve). In the adaptation phase, participants were divided into four groups, each with a different target training direction (lower left, lower right, upper right, or upper left), and visual rotation was gradually imposed on the tip-movement direction. Irrespective of the target direction, the adaptation pattern of the tipmovement and stick-tilt followed with the baseline relationship.

We also thank you for your comment about studying the existing redundant systems. We can understand the reviewer's concern about the usefulness of our task, but we believe that we have proposed the novel framework for motor adaptation in the redundant system. The future studies will be able to clarify how the knowledge gained from our task can be generally applied to understand the control and learning of the redundant system.

Reviewer #3 (Recommendations For The Authors):

Line 49: replace “uniquely” with primarily. A number of features of the task setup could affect the joint angles, from if/how the arm is supported, whether the wrist is fixed, alignment of the target in relation to the midline of the participant, duration of the task, and whether fatigue is an issue, etc. Your statement relates to fixed limb lengths of a participant, rather than standard reaching tasks as a whole. Not to mention the degree of inter- and intra-subject variability that does exist in point-to-point reaching tasks.

Thank you for your helpful point. We have replaced “uniquely” with “primarily”. (Line 49).

Line 72: the cursor is not an end-effector - it represents the end-effector.

We have changed the expression as “the perturbation to the cursor representing the position of the end-effector (Line 72).

Lines 73 – 78: it would benefit the authors to consider the role of intersegmental dynamics.

Thank you for your suggestion. We are not sure if we understand this suggestion correctly, but we interpret that this suggestion to mean that the end-effector perturbation can be implemented by using the perturbation that considers the intersegmental dynamics. However, the implementation is not so straightforward, and the panels in Figure 1j,k are only conceptual for the end-effector irrelevant perturbation. Therefore, we have not described the contribution of intersegmental dynamics here.

Lines 90 – 92: “cannot” should be “did not”, as the studies being referenced are already completed. This statement should be further unpacked to explain what they did do, and how that does not meet the requirement of redundancy in movement patterns.

We have changed “cannot” to “did not” (Line 91). We have also added the description of what the previous studies had demonstrated (Line 88-90).

Figure text could be enlarged for easier viewing.

We have enlarged texts in all figures.

Lines 41 - 47: Interesting selection of supporting references. For the introduction of a novel environment, I would recommend adding the support of Shadmehr and MussaIvaldi 1994.

Thank you for your suggestion. We have added Shadmehr and Mussa-Ivaldi 1994 as a reference (Line 45).

Line 49: “this task” is vague - the above references relate to a number of different tasks. For example, the authors could replace it with a reaching task involving an end-point robot.

Thank you very much for your suggestion. As per the suggestion by Reviewer #1, we have changed this to “such a planar arm-reaching task” (Line 49).

Line 60: “hypothetical limb with three joints” - in Figure 1a, the human subject, holding the handle of a robotic manipulandum does have flexibility around the wrist.

Previous studies using planar arm-reaching task have constrained the wrist joint (e.g., Flash & Hogan, 1985; Gordon et al., 1994; Nozaki et al., 2006). We tried to emphasize this point as “participants manipulate a visual cursor with their hands primarily by moving their shoulder and elbow joints” (Line 42). In the revised manuscript, we have also emphasized this point in the legend of Figure 1a.

Lines 93-108: this paragraph could be cleaned up more clearly stating that while the use of task-irrelevant perturbations has been used in the domain of reaching tasks, the focus of these tasks has not been specifically to address “In our task, we aim to exploit this feature by doing”

Thank you very much for your helpful comments. To make this paragraph clear, we have modified some sentences (Line 100-104).

Line 109: “coordinates to adapt” is redundant.

We have changed this to “adapts” (Line 110).

Lines 109-112: these sentences could be combined to have better flow.

Thank you very much for your valuable suggestion. We have combined these two sentences for the better flow (Line 110-112).

Line 113-114: consider rewording - “This is a redundant task because ...” to something like “Redundancy in the task is achieved by acknowledging that ....“.

We have changed the expression according to the reviewer’s suggestion (Line 114).

Line 118: Consider changing “changes” to “makes use of”.

We have changed the expression (Line 119).

Lines 346 - 348: grammar and clarity - “This redundant motor task enables the investigation of adaptation patterns in the redundant system following the introduction of perturbations that are either end-effector relevant, end-effector irrelevant, or both.“.

Thank you very much again for your helpful suggestion of English expression. We have adopted the sentence you suggested (Line 354-356).

Version 2

Abstract

Abstract

The remarkable ability of the motor system to adapt to novel environments has traditionally been investigated using kinematically non-redundant tasks, such as planar reaching movements. This limitation prevents the study of how the motor system achieves adaptation by altering the movement patterns of our redundant body. To address this issue, we developed a redundant motor task in which participants reached for targets with the tip of a virtual stick held with both hands. Despite the redundancy of the task, participants consistently employed a stereotypical strategy of flexibly changing the tilt angle of the stick depending on the direction of tip movement. Thus, this baseline relationship between tip-movement direction and stick-tilt angle constrained both the physical and visual movement patterns of the redundant system. Our task allowed us to systematically investigate how the motor system implicitly changed both the tip-movement direction and the stick-tilt angle in response to imposed visual perturbations. Both types of perturbations, whether directly affecting the task (tip-movement direction) or not (stick-tilt angle around the tip), drove adaptation, and the patterns of implicit adaptation were guided by the baseline relationship. Consequently, tip-movement adaptation was associated with changes in stick-tilt angle, and intriguingly, even seemingly ignorable stick-tilt perturbations significantly influenced tip-movement adaptation, leading to tip-movement direction errors. These findings provide a new understanding that the baseline relationship plays a crucial role not only in how the motor system controls movement of the redundant system, but also in how it implicitly adapts to modify movement patterns.

Introduction

If a movement error occurs while reaching a target, the motor system automatically and implicitly updates the motor command for the next movement to reduce the error. This remarkable motor learning ability contributes not only to maintaining accurate movements in the presence of noise but also to flexibly adapting the movement to new environments. In the standard planar arm-reaching task, which has been widely used to study the mechanisms of motor learning, participants manipulate a visual cursor with their hands primarily by moving their shoulder and elbow joints (Flash & Hogan, 1985; Gordon et al., 1994) (Fig. 1a). The arm-reaching task has contributed to uncovering many important properties of motor adaptation when a novel environment is introduced (Shadmehr & Mussa-Ivaldi, 1994; Thoroughman & Shadmehr, 1999; Mazzoni & Krakauer, 2006; Taylor & Ivry, 2011; Takiyama et al., 2015; Hayashi et al., 2016).

Kinematic redundancy and adaptation patterns.

(a) In the ordinary planar arm-reaching task which primarily uses elbow and shoulder joints, there is no redundancy between the task goal (xc, yc) and the joint angles (θ1, θ2). (b) The movement pattern at each target is determined by the joint angles. (c) The adaptation patterns to a visual rotation are also determined. (d) This adapted joint angle pattern should be identical to that of voluntary reaching in the adapted direction. (e, f) In the case of planar reaching movement by a hypothetical limb with three joints (θ1, θ2, θ3), the cursor can reach the target in a large number of patterns. (g) Given that the motor system solves the task redundancy by optimization, we should observe a stereotypical pattern of joint angles according to the targets. (h, i) The question is whether, when a visual rotation is imposed on the cursor (end-effector relevant perturbation), the motor system adopts the same joint movement pattern as when voluntarily aiming in the adapted direction (left) or a new joint movement pattern (right). (j) The joint angles can be perturbed while the cursor position remains unchanged (end-effector irrelevant perturbation). (k) This perturbation can be ignored because it does not affect the cursor (left). Alternatively, the joint angles may be corrected (right).

Notably, there is no redundancy between hand position and joint angles in such a planar arm-reaching task. The joint angles are primarily determined by the target to be reached (Fig. 1b). Because of this non-redundancy, when the motor system adapts the reaching movement to the visual rotation imposed on the cursor (Fig. 1c, left), the adapted pattern of joint angles is also determined (Fig. 1c, right). Naturally, the adapted pattern of joint angles should be identical to the pattern when reaching in the adapted direction (Fig. 1d). However, our body systems for performing practical motor tasks generally have more degrees of freedom than necessary to accomplish the tasks (Bernstein, 1996; Singh et al., 2016). For example, in the golf swing, there are countless patterns of whole-body configurations to achieve identical clubhead positions at ball contact.

The problem arising from such a redundant structure can be illustrated by considering the reaching movement of a hypothetical limb with three joints (Fig. 1e). In this case, there are a large number of possible joint angle patterns for the same hand position (Fig. 1f). However, given that the motor system solves task redundancy by minimizing motor cost (Collins, 1995; O’Sullivan et al., 2009) or movement error (White & Diedrichsen, 2010; Zhang et al., 2018), we should observe a stereotypical pattern of joint angles according to the targets (Fig. 1g). The problem is how the motor system changes the direction of cursor movement to implicitly adapt to the visual rotation imposed on the cursor (Fig. 1h). Are the joint angles after adaptation identical to those observed when aiming in the same adapted direction before adaptation? (Fig. 1i, left). Alternatively, does the motor system adopt a new joint movement pattern? (Fig. 1i, right).

It has not been thoroughly investigated how the motor system coordinates to change the redundant body system to implicitly adapt to the perturbation to the cursor representing the position of the end-effector.

Redundancy poses another type of problem for implicit movement adaptation. If the joint angles are perturbed while the hand position remains unchanged (Fig. 1j), how does the motor system respond to the perturbation? The motor system may simply ignore the perturbation because it does not affect the cursor position (Fig. 1k, left). Alternatively, the motor system may attempt to correct the joint movements (Fig. 1k, right). Conventionally, this type of problem has been investigated in the context of how the motor system responds to the movement variability that affects or does not affect task performance (i.e., task-relevant or task-irrelevant variability). Previous studies have consistently shown that the motor system actively corrects for task-relevant variability while allowing task-irrelevant variability (Mosier et al., 2005; Latash, 2012; van Beers et al., 2013; Sternad, 2018; Scholtz & Schoner, 1999; Scholtz et al., 2000). Optimal feedback control theory suggests that the strategies based on such a “minimal intervention principle” are optimal in terms of achieving task goals (e.g., movement accuracy) (Todorov & Jordan, 2002; Todorov, 2004; Scott, 2004; Diedrichsen, 2007; Valero-Cuevas et al., 2009; Ronsse et al., 2010; Sternad et al., 2011; Selgrade & Chang, 2015). These studies demonstrated that the motor system does not intervene movement change or variability as long as the task goal is achieved. Thus, the motor system should ignore the perturbations to joint angles that do not affect the task goal (Fig 1k, left). However, these studies did not directly test whether the redundant body movement pattern is also unchanged when the end-effector movement remains unchanged.

Conversely, it has also been reported that the motor system adapts to change the movement in response to the task-irrelevant perturbation (Wolpert et al., 1995; Flanagan & Rao, 1995; Morehead et al., 2017; Diedrichsen et al., 2010; Schaefer et al., 2012; Franklin et al., 2016; Kim et al., 2019). For example, when the cursor trajectory was curved only in the middle of the reaching movement, the hand movement was adaptively modified even though the perturbation did not cause a reaching error at the target (Wolpert et al., 1995; Flanagan & Rao, 1995). It should be noted that these previous studies created the task-irrelevant condition by applying the perturbation to the end-effector. In contrast, the present study focuses on how the motor system responds to the perturbation of the redundant body the end-effector remains unaffected (Figs. 1i and 1k). In light of this, in this study, we use the terms “end-effector relevant perturbation” (e.g., Fig. 1h) and “end-effector irrelevant perturbation” (e.g., Fig. 1j) when appropriate, instead of the more commonly used terms “task-relevant perturbation” and “task-irrelevant perturbation”. The current study aims to develop a novel task that creates end-effector relevant and end-effector irrelevant perturbations.

To investigate the problem of how the motor system adapts the redundant body system to these perturbations, this study introduces a novel bimanual stick manipulation task specifically designed to address this issue (Fig. 2a). Participants manipulated a virtual stick held with both hands to reach visual targets by the tip of the stick. Redundancy in the task is achieved by acknowledging that the same tip position can be achieved with different stick angles (Fig. 2b). Importantly, this task allowed us to visually perturb the tip position (i.e., end-effector relevant perturbation) and the stick angle (i.e., end-effector irrelevant perturbation) either separately or simultaneously. Using this approach, we can directly investigate how the motor system makes use of the redundant body system when adapting to these two types of perturbations.

Bimanual stick manipulation task.

(a) Participants held handles of a manipulandum with both hands and moved the right tip of a virtual stick on the monitor. (b) In this task, there are countless movement patterns to move the tip to the target. Two representative strategies are shown: parallel translation (top) and tilting the stick (bottom). (c) During the baseline phase, participants were instructed to move the tip from a starting point to a target that appeared in one of nine directions. (d) Movement trajectories (top) and velocity profiles (bottom) of the left hand, right hand, and tip averaged across participants. Each color corresponds to the target direction shown in c. The shaded areas represent the SEM across participants. (e) The time-course of the stick-tilt angle shows that participants moved the tip while tilting the stick. The shaded areas indicate the SEM. (f) Trial-to-trial changes in stick-tilt angle at the peak velocity of the tip. The strategy of tilting the stick was observed from the beginning of the experiment (t-test compared with zero, |t(71)| > 6.4, Bonferroni corrected P < 0.001). The shaded areas indicate the SEM.

Results

Movement strategy to manipulate the stick

A total of 72 healthy participants took part in the study (Experiments 1–3). They were asked to manipulate a virtual stick with both hands (length = 40 cm; distance between hands = 15 cm) to move the right tip of the stick towards the targets (Fig. 2a). This task is redundant because the same target can be reached in countless movement patterns. The participants could move both hands in parallel without tilting the stick (Fig. 2b, top). Alternatively, they could perform the task by tilting the stick (Fig. 2b, bottom). During the first part of the task (baseline phase: 360 trials) in all the experiments (Experiments 1–3), the participants moved the tip from a starting point to a target (movement distance = 10 cm) that appeared in one of nine directions (0°: horizontal direction, ±10°, ±20°, ±30°, and ±40°) (Fig. 2c).

The trajectories of the tip and both hands appeared to be almost straight, but both hands were not moved in parallel, except when aiming at the 0° target (Fig. 2d, top). The velocity profiles of the hands and tip were bell-shaped (Fig. 2d, bottom), similar to a normal unimanual reaching movement (Flash & Hogan, 1985; Gordon et al., 1994). The time-course changes in the stick-tilt angle relative to the initial horizontal tilt angle showed that the participants moved the tip while slightly tilting the stick, except when aiming at the 0° target (Fig. 2e). This tendency was observed from the beginning of the experiment (t-test compared with 0°, |t(71)| > 6.4, Bonferroni corrected P < 0.001; Fig. 2f).

Movement strategies to manipulate the stick for individuals are illustrated on the plane expressed by two variables: tip-movement direction relative to a horizontal direction (TMD) and the stick-tilt angle relative to the initial horizontal state (STA) (Fig. 3a). Although each participant could have performed the task differently (Fig. 2b), all participants showed stereotypical movement patterns: The stick-tilt angle was significantly modulated by the tip-movement direction (t-test compared with 0°, |t(71)| > 5.5, Bonferroni corrected P < 0.01; Fig. 3a, inset). This monotonically increasing curve, which will be referred to as the “baseline TMD–STA relationship”, reflects a constraint between the direction of tip movement and the angle of stick tilt. The relationship showed that clockwise (CW) tip movement direction was accompanied by CW stick-tilt, and counterclockwise (CCW) tip-movement direction was accompanied by CCW stick-tilt. This constraint of movement patterns was likely created by minimizing the movement distance of the hands. If both hands are moved in parallel, the total distance between the initial and final positions of the hands should be 20 cm (10 cm for each) regardless of the targets; however, the experimentally observed distance was reduced by tilting the stick (Fig. 3b). The average distance was a near- minimal value (versus parallel translation: Cohen’s d = 1.914, versus theoretical minimum: Cohen’s d = 0.334; Fig. 3c).

Relationship between the tip-movement direction and stick-tilt angle.

(a) A relationship between each individual’s tip-movement direction and stick-tilt angle, when the tip speed peaked is represented as the thin gray lines drawn with colored plots. The averaged relationship is plotted as a bold gray line (baseline TMD–STA relationship) with the SD (crosses) and 95% confidence intervals (ellipses). The inset shows the stick-tilt angles at peak velocity and movement offset (error bars: SD, *: t-test compared with 0°, Bonferroni-corrected P < 0.01). The monotonically increasing curve indicates the stereotypical strategy of manipulating the stick (diagrams on the right side). (b) Sum of the distances between the initial and final positions of both hands (i.e., the sum of the lengths of pink arrows in Fig. 2b). The dotted and dashed gray lines indicate the distances when both hands are moved in parallel and the mathematically derived minimum distances by tilting the stick, respectively. The error bars denote the SD. (c) The distance averaged across target directions was closer to the minimum value (Cohen’s d = 1.914) than the parallel translation distance (Cohen’s d = 0.334). The error bar indicates the SD across participants.

Experiment 1: Motor adaptation pattern to the end-effector relevant perturbation

During the second part of the experiment (adaptation phase: 240 trials), the target appeared only in the 0° direction, and three types of visual perturbations were introduced. Most importantly, this task enabled the introduction of end-effector relevant perturbation (Experiment 1), or end-effector irrelevant perturbation (Experiment 2), or a combination of both types of perturbations (Experiment 3). In Experiment 1A (N = 19), the tip-movement direction was visually rotated in the CCW direction around the starting position (tip perturbation), whereas the stick-tilt angle remained unchanged (Fig. 4a). Thus, this perturbation produced a positional dissociation between vision (i.e., the stick) and proprioception (i.e., the hands). The amount of tip directional rotation was gradually increased by 1° per a trial up to 30° so that the participants did not notice the presence of perturbation.

Experiment 1: End-effector relevant visual perturbations.

(a) During the adaptation phase of Experiment 1, the tip-movement direction was rotated by 30° around the starting point (tip perturbation). The dotted and solid black arrows indicate the intended and actual tip trajectory, respectively. (b) The changes in the movement pattern are expressed as the state shift on the plane represented by mapping between tip-movement direction and stick-tilt angle (left: visual space, right: physical space). The dashed red lines denote the task goal (i.e., a target direction). On the visual plane, the perturbation (a blue arrow) shifts the movement pattern. On the physical plane, the task goal is shifted. (c) The participants could correct the tip error without (pattern i, top) or by (pattern ii, bottom) altering the stick-tilt angle. (d) The adaptive strategies are expressed as the shift on the visual plane (left) and the physical plane (right). The visual error caused by the perturbation could be compensated by different adaptation patterns (blue arrows): Representative patterns are expressed as bold arrows (patterns i and ii).

Figure 4b illustrates how this visual perturbation changes the states of stick movement in visual and physical spaces represented by mapping between the tip- movement direction and stick-tilt angle. The movement in the visual space was the one visible to the participants, whereas the movement in the physical space was the actual hand movement, which was invisible to them. On a plane representing visual movement patterns (Fig. 4b, left), the visual perturbation to the tip can be expressed as a shift of the state along the x-axis. The error of the tip-movement direction in the visual space needs to be compensated by changing the movement of both hands. On a plane representing physical movement patterns (Fig. 4b, right), this requirement can be expressed as a shift of the task goal in the direction opposite to the visual perturbation.

Owing to the redundant nature of this task, participants could adapt to change the visual tip movement using different strategies. For example, they could change the tip movement without changing the stick-tilt while keeping the strategy of moving both hands in parallel (Fig. 4c, top). Alternatively, they might adapt by changing the stick tilt as if they were aiming at a -30° target (Fig. 4c, bottom). These strategies can be expressed as a state shift towards the new task goal line along the x-axis or along the baseline TMD–STA relationship in the physical plane (Fig. 4d).

The experimental results demonstrated that the participants adapted to the tip perturbation by tilting the stick (Fig. S1). On the visual plane (Fig. 5a), after the adaptation, the tip-movement direction remained at the baseline level (paired t-test, t(18) = -1.320, P = 0.203), while the stick was tilted significantly more in the CW direction than when aiming at the 0° target in the baseline phase (paired t-test, t(18) = 10.003, P < 0.01). Thus, the tip perturbation was compensated by adaptation, but the adaptation produced an abnormal visual stick-tilt. This result is likely to support the idea of the minimal intervention principle because participants ignored this visual stick- tilt which did not directly influence the task.

Experiment 1: Adaptation patterns to gradually imposed end-effector relevant perturbations.

(a) In Experiment1A, the adaptation trajectory (a thin blue line) on the visual plane gradually deviated from a baseline pattern (a gray circle) mainly along the y-axis. The tip-movement direction remained at the baseline level (paired t- test, t(18) = -1.320, P = 0.203). In contrast, the stick-tilt angle after the adaptation (a blue circle) significantly differed from that in the baseline pattern when aiming at the 0° target (a gray circle) (paired t-tests, t(18) = 10.003, P < 0.001), indicating that the participants saw the abnormal visual stick-tilt. The plots indicate mean ± SD across participants. (b) On the physical plane, the adaptation trajectory followed the baseline TMD–STA relationship. Although the difference was small but significant for the tip- movement direction compared with the baseline pattern when aiming at the -30° target (paired t-test, t(18) = -10.053, P < 0.01), there was no significant difference in the stick-tilt angle (paired t-test, t(18) = -1.081, P = 0.294). (c) The area indicated by the squared outline in b is enlarged. The adaptive process can be divided into nine trial segments (trials 1–5, 6–10, 11–15, 16–20, 21–25, 26–30, 30–120, 121–210, and 211–240). (d) The motor system implicitly achieved adaptation to the tip perturbation as if it had voluntarily aimed at the -30° target. (e) Participants who showed a larger stick-tilt when aiming at the -30° target during the baseline phase also exhibited a larger stick-tilt in the adapted pattern (R = 0.601, P < 0.007).

On the physical plane, adaptation appeared to progress along the baseline TMD–STA relationship throughout the adaptation phase (Figs. 5b and 5c). The implicitly adapted tip-movement direction and stick-tilt angle were similar to those when voluntarily aiming at the -30° target in the baseline phase. Although the difference was small but significant for the tip-movement direction (paired t-test, t(18) = -10.053, P < 0.01), there was no significant difference in the stick-tilt angle (paired t-test, t(18) = - 1.081, P = 0.294). Thus, the motor system implicitly achieved adaptation to the tip perturbation as if it had voluntarily aimed at the -30° target (i.e., pattern ii in Figs. 4c and 4d) (Fig. 5d). This adaptation characteristic was also supported by the observation that the participants who tilted their stick more when aiming at the -30° target in the baseline phase exhibited a greater stick- tilt in the adapted movement patterns (R = 0.601, P = 0.007; Fig. 5e). In Experiment 1B (N = 13), we imposed the 30° tip perturbation abruptly to allow participants to use explicit strategy, and obtained nearly identical results (Figs. S2 and S3).

As described above, the adapted tip-movement direction on the physical plane was significantly different from the tip-movement direction when aiming at the -30° target in the baseline phase (Fig. 5b). In contrast, such significant difference was not observed for Experiment 1B (Fig. S3). Why was such a difference only in Experiment 1A? We speculated that this difference was caused by the implicit adaptation to correct the abnormal visual stick-tilt angle. As shown in Fig. 5a, after the adaptation, the stick was visually tilted in the CW direction. If the physical movement pattern is constrained by the baseline TMD–STA relationship, the CCW correction of the stick-tilt angle should be accompanied by the CCW change in tip-movement, which would interfere with the adaptation to the tip perturbation.

Experiment 2: Motor adaptation pattern to end-effector irrelevant perturbation

Thus, it is still possible that the motor system does not necessarily ignore, but corrects for, the stick-tilt error that does not directly influence the tip position. Experiment 2 (N = 20) was designed to directly test this possibility by imposing a stick-tilt perturbation around the right tip (i.e., end-effector irrelevant perturbation) (Fig. 6a). As the tip moved from the starting point, the stick started to tilt around the tip up to ±6° when the tip reached the target (the tilt amount was proportional to the tip’s distance from the starting point). The ±6° tilt perturbation was chosen according to the stick tilt at the movement offset when reaching the ±30° target for the baseline phase (see Fig. 3a). Ten participants participated in the CCW stick-tilt perturbation condition (E2CCW group: 6°), and the remaining ten participants took part in the CW stick-tilt perturbation condition (E2CW group: -6°). Because the amount was small, the participants did not notice the presence of perturbation throughout the experiment. Since this perturbation changed only the stick-tilt angle, it is expressed as a state shift along the y-axis on the visual plane (Fig. 6b). According to the minimal intervention principle, the motor system does not need to change anything because the stick-tilt perturbation does not affect the tip position (i.e., the task goal) (Figs. 6c and 6d, pattern i). However, if the motor system is actively compensating for the stick-tilt perturbation, it should change the stick-tilt angle (Figs. 6c and 6d, pattern ii).

Experiment 2: End-effector irrelevant visual perturbations.

(a) During the adaptation phase of Experiment 2, the stick-tilt angle was rotated around the tip. The perturbation of ±6° was imposed from the first trial. Participants received either the counterclockwise perturbation (green, E2CCW group) or the clockwise perturbation (orange, E2CW group). (b) On the visual plane, the perturbation is expressed as a vertical shift of the state along the task goal (dashed red line). On the physical plane, the perturbation does not influence the task goal. Note that, in both planes, the states are on the task goal lines. The formats are the same as in Fig. 4b. (c) Participants may not change the movement pattern because the perturbation does not affect the tip position (pattern i, top). Alternatively, the perturbation can lead to a stick-tilt correction (pattern ii, bottom). (d) These strategies are expressed as a shift on the plane. Note that pattern i means that there is no change in the movement pattern.

Interestingly, the participants corrected the stick-tilt angle (Fig. S4). On the visual plane, the stick-tilt angles remained different from those in the baseline phase (Fig. 7a) (two-way (direction of stick-tilt perturbation (i.e., groups), trial phase (first 30 trials or last 30 trials)) repeated-measures ANOVA, a significant main effect of groups: F(1,18) = 54.457, P < 0.001; no effect of trial phase: F(1,18) = 2.554, P = 0.127; no interaction: F(1,18) = 1.613, P = 0.220). Thus, the adaptive changes did not fully compensate for the perturbation-induced stick-tilt errors. However, the data on the physical plane clearly demonstrated that participants implicitly changed the stick-tilt angle in the direction opposite to that of the perturbation from the early adaptation phase (Fig. 7b): Without the correction, the data of the E2CCW and E2CW groups should have largely overlapped. However, with respect to stick-tilt angle, two-way repeated-measures ANOVAs revealed the main effect of stick-tilt perturbation direction (F(1, 18) = 12.483, P < 0.01) but not the main effect of trial phase (F(1, 18) = 3.331, P = 0.085). The interaction between these factors was not significant (F(1, 18) = 1.362, P = 0.258).

Experiment 2: Adaptation patterns to end-effector irrelevant visual perturbations.

(a) Movement patterns are shown by green (E2CCW group) and orange (E2CW group) plots on the visual plane. The open and filled plots represent the movement patterns in the first and last 30 trials, respectively, during the adaptation phase (mean ± SD, across participants). (b) On the physical plane, the separated movement patterns between both groups showed that participants corrected the end-effector irrelevant stick-tilt errors. (c) The enlarged plot of the squared area in b shows that end-effector relevant tip-movement directional errors accompanied the adaptation. These effects persisted until the end of the adaptation phase (two-way repeated-measures ANOVAs). (d) Schematics of how the stick-tilt perturbation altered the movement patterns. (e, f) How participants tilted the stick during the baseline phase was negatively correlated with the stick-tilt angle and the tip-movement direction. To control for the individual variability, the baseline stick-tilt angle or tip-movement direction when aiming at the 0° target was subtracted from each value.

Unexpectedly, this stick-tilt correction was accompanied by undesirable tip- movement directional errors (Fig. 7c). With respect to the tip-movement direction, two- way repeated-measures ANOVAs revealed the main effect of the direction of stick-tilt perturbation (F(1, 18) = 5.104, P = 0.037) but not the main effect of trial phase (F(1, 18) = 1.288, P = 0.271) (the interaction between these factors was not significant: F(1, 18) = 0.804, P = 0.382). Notably, these seemingly irrational adaptation patterns were consistent with the patterns predicted from the baseline physical TMD–STA relationship: CCW (CW) stick-tilt adaptation to the CW (CCW) stick-tilt perturbation slightly shifted the tip-movement direction in the CCW (CW) direction (Fig. 7d).

We also examined how the amount of adaptation depended on the amount of stick-tilt when voluntarily aiming at the ±30° target in the baseline phase. The 6° stick- tilt perturbation should be insignificant for participants who tilt the stick at a larger angle; thus, this perturbation tends to be ignored. Conversely, for participants who did not tilt the stick, a 6° stick-tilt perturbation should be significant and must be corrected. Figures 7e and 7f show how the amount of stick-tilt correction (Fig. 7e) and tip-movement directional change (Fig. 7f) depended on the amount of stick-tilt angle when aim at the ±30° target in the baseline phase. As predicted, we observed a negative correlation (stick-tilt correction: R = -0.572, P < 0.01; tip-movement directional change: R = -0.440, P = 0.052), although the tip-movement direction change was not significant.

Taken together, the results of Experiment 1 and Experiment 2 suggest that the implicit adaptation in the redundant stick-manipulating task is achieved by the following two mechanisms: First, the motor system attempts to correct for any perturbation in visual feedback, regardless of whether it is end-effector relevant or end-effector irrelevant. Second, the physical pattern of movement correction is constrained by the baseline TMD–STA relationship.

Experiment 3: Interaction between end-effector relevant and end-effector irrelevant perturbations

Experiment 3 (N = 20) was designed to further investigate the interaction between adaptations to end-effector relevant and end-effector irrelevant perturbations by applying both simultaneously (Fig. 8a). We hypothesized that adaptation to the tip perturbation would be influenced by the direction of the simultaneously applied stick- tilt perturbation. Specifically, the CCW tip perturbation was combined with the CCW stick-tilt perturbation (E3CCW group: N = 10) or with the CW stick-tilt perturbation (E3CW group: N = 10). These visual perturbation patterns can be expressed as state shifts in the visual plane (green and orange arrows for E3CCW and E3CW groups, respectively, in Fig. 8b). On the physical plane, the correction pattern required for the E3CCW group (Fig. 8c, top) is aligned with the baseline TMD–STA relationship (green arrow in Fig. 8d), from which we can predict that adaptation should proceed smoothly. In contrast, the correction pattern required for the E3CW group (Fig. 8c, bottom) is orthogonal to the baseline TMD–STA relationship (orange arrow in Fig. 8d), which may prevent smooth adaptation. We tested these predictions by simultaneously imposing the tip perturbation which was the same as the perturbation used in Experiment 1A and the stick-tilt perturbation which was the same as the perturbation used in Experiment 2. Again, the participants were unaware of the perturbations throughout the experiment.

Experiment 3: Interaction between end-effector relevant and end- effector irrelevant visual perturbations.

(a) During the adaptation phase of Experiment 3, both the CCW tip perturbation and the CCW or CW stick-tilt perturbation were introduced. The participants were assigned to the CCW stick-tilt rotation condition (green, E3CCW group) or the CW stick-tilt rotation condition (orange, E3CW group). (b) These visual perturbations can be expressed as state shifts on the visual plane (green and orange arrows for E3CCW and E3CW groups, respectively). (c) The adaptation pattern to reduce visual errors (top: E3CCW group, bottom: E3CW group). (d) On the physical plane, the correction pattern required for the E3CCW group is aligned with the baseline TMD–STA relationship (green arrow), whereas the correction pattern required for the E3CW group is orthogonal to the baseline TMD–STA relationship (orange arrow).

Contrary to the predictions, the difference in stick-tilt perturbation directions did not affect the final adapted movement patterns (Fig. S5): As shown in the visual plane (Fig. 9a), the difference in stick-tilt angle remained unchanged (paired t-test, t(18) = 6.920, P < 0.001). Similarly, in the physical plane (Fig. 9b), there were no significant differences in movement pattern between the two groups (tip-movement direction: paired t-test, t(18) = -0.479, P = 0.638, stick-tilt angle: paired t-test, t(18) = -0.386, P = 0.704). However, a trial-by-trial variability of the movement pattern in the latter adaptation phase (trial# 121–240) showed a slight difference between groups (t-test, tip-movement direction: t(18) = -2.046, P = 0.056, stick-tilt angle, t(18) = -3.235, P = 0.005), suggesting that the adaptation processes were implicitly affected by combination of perturbations (Fig. 9a, inset). In partial agreement with our predictions, two groups showed different adaptation processes: trial-dependent trajectories to the final adapted pattern were clearly separated for both groups (Fig. 9c). Indeed, a repeated-measures multivariate analysis of variance (MANOVA) showed a significant effect of group (Pillai’s Trace F(8, 162) = 0.155, P < 0.001) and trial segment (Pillai’s Trace F(8, 162) = 0.719, P < 0.001); however, no significant interaction was observed (Pillai’s Trace F(8, 162) = 0.039, P = 0.982). As a result, the adaptation of the tip movement was more delayed in the E3CW group (trial segment 2: t-test, t(18) = -3.684, Bonferroni-corrected P < 0.05; Fig. 9d).

Experiment 3: Adaptation patterns to end-effector relevant and end- effector irrelevant visual perturbations.

(a) Adaptation patterns on the visual plane are shown (same formats as Fig. 5a). A green (orange) plot represents the adapted movement pattern in the E3CCW group (E3CW group). The tip-movement direction did not significantly differ (t-test, t(18) = -0.473, P = 0.642), while the stick-tilt angle remained a significant difference between groups (t-test, t(18) = 6.920, P < 0.001). A trial-by-trial variability of the movement pattern in the latter adaptation phase (trial# 121–240) showed a difference between groups (t-test, tip-movement direction: t(18) = -2.046, P = 0.056, stick-tilt angle, t(18) = -3.235, P = 0.005) (b) Similarly, the adapted movement patterns on the physical plane did not show a significant difference (t-test, tip-movement direction: t(18) = -0.479, P = 0.638, stick-tilt angle: t(18) = -0.386, P = 0.704). (c) However, the adaptive processes appeared to be different. The movement patterns of the two groups are compared using a repeated-measures MANOVA. A multivariate analysis revealed a significant effect of group (Pillai’s trace F(8, 162) = 0.155, P < 0.001) and trial segment (Pillai’s trace F(8, 162) = 0.719, P < 0.001); no significant interaction can be observed (Pillai’s trace F(8, 162) = 0.039, P = 0.982). The definition of trial segments is the same as that shown in Fig. 5c. (d) The learning curves for the tip- movement. The adaptation in the E3CW group was always delayed but showed a significant difference only during the early phase (trial segment 2: t-test, t(18) = -3.684, Bonferroni-corrected P < 0.05).

Discussion

Our bodies have more degrees of freedom than necessary to achieve a relatively low- dimensional task goal. Understanding how the motor system resolves this redundancy remains a challenging problem (Bernstein et al., 1996). In this study, we investigated the problems of motor control and adaptation specific to kinematically redundant systems. The first problem is how the motor system coordinates the redundant body movements to control the end-effector (Fig. 1f). The second problem is how the motor system coordinates to modify redundant body movements when implicitly adapting to the perturbation imposed on the end-effector (Fig. 1i). The third problem is how the motor system responds to the perturbation to the redundant dimension that does not directly affect the end-effector (Fig. 1k). These two types of perturbations have been conventionally referred to as “task-relevant” and “task-irrelevant” perturbations. However, it should be noted that previous studies investigating the effect of task- irrelevant perturbations have applied perturbations to the end-effector (e.g., cursor) without affecting task performance (Wolpert et al., 1995; Flanagan & Rao, 1995; Diedrichsen et al., 2010; Schaefer et al., 2012; Franklin et al., 2016; Kim et al., 2019). In contrast, the present study focused on investigating the effect of perturbations on the redundant body while leaving the end-effector unaffected (see Figure 1k). To clearly emphasize these differences, we specifically used the terms ’’end-effector relevant perturbation’’ and ’’end-effector irrelevant perturbation.’’

To investigate these issues, we designed a novel stick manipulation reaching task. This redundant motor task enables the investigation of adaptation patterns in the redundant system following the introduction of perturbations that are either end-effector relevant, end-effector irrelevant, or both. Because our experiments were conducted with a specific configuration (stick length, hand distance, target location etc), we must be cautious about the generalizability of the results to other configurations, but we obtained the following 4 main results. First, we showed that participants adopted a stereotyped strategy to reach the target with the stick tip, and that the movement pattern (stick tip movement and stick tilt angle) was constrained by a baseline TMD– STA relationship. Second, when only end-effector relevant perturbations were imposed on the tip, the motor system attempted to modify the physical movement patterns along the baseline TMD–STA relationship to achieve the task goal (Experiment 1). Third, even when perturbations were applied only to the end-effector irrelevant dimension (i.e., the stick-tilt angle), the motor system still adapted to the end-effector irrelevant perturbation and this adaptation was accompanied by an end-effector relevant error (Experiment 2). Finally, when both types of perturbations were imposed simultaneously, the adaptation processes were influenced depending on the combination of perturbations (Experiment 3).

Kinematic relationship between end-effector relevant and end-effector irrelevant dimensions

In our stick manipulation task, participants were asked to reach the tip of the stick to the targets. Participants can achieve the reaching task in different ways. For example, they can use a simple strategy by moving both hands in the same direction (Fig. 2b, top) or a more complicated strategy by tilting the stick (Fig. 2b, bottom). Nevertheless, they showed stereotypical patterns of moving the stick tip by tilting the stick from the beginning of the task (Fig. 2f): Reaching the CW (or CCW) targets was accompanied by the CW (or CCW) stick tilt (Fig. 3a). Such a stereotypical movement patterns have been reported for the movements of the joints in the redundant 3-dimensional reaching movements (Lacquaniti & Soechting, 1982; Proietti et al., 2017). Notably, compared with the case where both hands moved in parallel, the movement distances of both hands to perform the task were significantly reduced to a near-optimal level (Figs. 3b and 3c), which contributed to the reduction of work required (Collins, 1995; O’Sullivan et al., 2009). As a result of this optimization process, tip-movement and stick-tilt were associated with each other by the baseline TMD–STA relationship (Fig. 3a). Physically, this relationship constrains how the motor system should move the stick. Visually, it tells the motor system how the stick movement should be seen.

The strength of the task to investigate the motor adaptation in redundant system

Our task has several advantages. First, the task has created redundancy in a natural way. Manipulating a stick with both hands is relatively common in everyday life (using a broom, golf or bat swing etc.). This familiarity with the task was reflected in Fig. 2f which shows that all participants started tilting the stick to reach the tip toward the targets from the beginning of the experiment. This characteristic is ideal for investigating how this inherent movement pattern influences motor adaptation. Second, this task allows us to investigate how the perturbation to the tip-movement direction leads to a change in both the tip-movement direction itself and the stick-tilt angle. It has not been investigated how the motor system implicitly achieves adaptation to the end-effector perturbation (i.e., the end-effector relevant perturbation) by changing the movement patterns of the redundant body system. Third, this task can create a novel condition in which the perturbations are imposed on the redundant system while the tip position remains unchanged (i.e., the end-effector irrelevant perturbation). No previous studies have investigated how the motor system implicitly responds to such a perturbation by altering (or not altering) the movement of the redundant body system.

Adaptation to end-effector relevant perturbation

When the end-effector relevant perturbation was implicitly imposed to the tip- movement direction (visual rotation by 30° CCW), the participants adapted to change the tip-movement direction in the opposite CW direction (Experiment 1), as reported in many previous studies using an ordinary reaching movement (Cunningham, 1989; Krakauer et al., 2000). Importantly, our task helped to investigate how such motor adaptations in the end-effector relevant dimension (tip-movement direction) are accompanied by the movement pattern changes in the end-effector irrelevant dimension (stick-tilt angle). Participants implicitly corrected the tip-movement direction by changing the stick tilt as if they were voluntarily aiming at the -30° target. This adaptation can be visualized as a state shift along the baseline TMD–STA relationship in the physical plane (Fig. 5b). The baseline TMD–STA relationship originally contributes to dictate how the motor system voluntarily moves the stick. It is likely that the motor system also uses this relationship as a scaffold to implicitly guide how to correct the tip-movement direction. Hayashi et al. (2016) have demonstrated that the adaptation of the standard reaching movement to a visual perturbation was significantly influenced by the baseline relationship between the movement and target directions (i.e., a visuomotor map). After distorting the visuomotor map so that the movement direction was more (or less) sensitively modulated by the target direction, the movement adaptation was enhanced (or reduced). The present finding that the baseline movement pattern is used for the implicit adaptation is consistent with this previous finding.

Notably, this adaptation pattern changed the visual feedback of the stick tilt. When reaching the 0° target in the baseline condition, the stick-tilt angle was almost 0° (Fig. 3a). However, after adaptation, the amount of stick-tilt was equivalent to that when aiming at the -30° target (Fig. 5b). Thus, the visually observed stick-tilt after adaptation was different from that observed in the baseline condition (Fig. 5a). Nevertheless, the adapted physical movement pattern persisted to follow the baseline TMD–STA relationship, suggesting that the motor system mainly paid attention to end-effector relevant tip-movement and ignored end-effector irrelevant stick-tilt angle. This adaptation pattern seems to be consistent with the minimal intervention principle.

However, a slight difference in the adaptation patterns between Experiment 1A and Experiment 1B implies a violation of the minimal intervention principle. In Experiment 1A, in which a 30° visual rotation was implicitly introduced by gradually increasing its amplitude, the final adapted tip-movement direction was slightly but significantly different from that when voluntarily aiming at the adapted direction (i.e., - 30° target) (Fig. 5b). In contrast, such residual difference was not observed in Experiment 1B, where participants could use an explicit aiming strategy (Fig. S1b). Visuomotor adaptation is often incomplete even after sufficient training, and residual errors are observed (Krakauer et al., 2000; Ethier et al., 2008; Malone et al., 2011; Vaswani et al., 2015). Although the presence of residual errors has been attributed to the imbalance between learning and forgetting, a recent study demonstrated that only the implicit learning system is responsible for producing residual errors by modulating error sensitivity with the consistency of errors in previous trials (Albert et al., 2021). We speculate that an alternative mechanism is involved in the generation of a residual difference in Experiment 1A. If the motor system does not ignore the abnormal stick- tilt but attempts to correct it, and this correction pattern is physically constrained by the baseline TMD–STA relationship, the stick-tilt correction should accompany the de- adaptation of the tip-movement direction and lead to the generation of a residual error. As discussed below, the results of Experiment 2 and Experiment 3 support this idea.

Adaptation to end-effector irrelevant perturbation

Our task can create a novel end-effector irrelevant stick-tilt perturbation and allows us to study how the motor system responds to it. According to the minimal intervention principle, the motor system should ignore stick-tilt perturbations because they do not influence task goal achievement (i.e., so called “task-irrelevant” perturbation). However, inconsistent with this idea, Experiment 2 clearly demonstrated that the motor system responded to the stick-tilt perturbations (Figs. 7b and 7c). This result indicates that the motor system predicts the visual feedback of an end-effector irrelevant stick-tilt angle when performing the task and tries to eliminate the dissociation between the observed and predicted stick-tilt angle (i.e., sensory prediction error). This idea was corroborated by the fact that participants who demonstrated a shallower TMD–STA relationship during the baseline phase showed a greater amount of stick-tilt correction (Fig. 7e). The ±6° stick-tilt perturbation should cause the larger prediction error for these participants.

One might consider that previous studies have already shown that the motor system responds to the task-irrelevant perturbations (Wolpert et al., 1995; Flanagan & Rao, 1995; Diedrichsen et al., 2010; Schaefer et al., 2012; Franklin et al., 2016; Kim et al., 2019). However, the task-irrelevant conditions implemented in these previous studies are quite different from the end-effector irrelevant condition in our study. These studies created the task-irrelevant condition by combining the perturbation of visual cursors with the manipulation of movement trajectory (Wolpert et al., 1995; Flanagan & Rao, 1995; Morehead et al., 2017), target/cursor shape (Diedrichsen et al., 2010; Schaefer et al., 2012; Kim et al., 2019), or applying the simultaneous equal perturbations to the target and cursor (Franklin et al., 2016). Therefore, these perturbations are not “end-effector irrelevant”. Exceptionally, in several previous studies, an end-effector irrelevant perturbation was implemented in a bimanual postural task to investigate postural reflex activity (Dimitriou et al., 2012; Omrani et al., 2013). However, to our knowledge, such an end-effector irrelevant perturbation has never been used to investigate the motor adaptation pattern in a redundant system. Furthermore, as discussed in the next section, our task design revealed a novel aspect of motor adaptation to the end-effector irrelevant perturbation.

Adaptation to end-effector irrelevant perturbation influences end-effector relevant performance

Unexpectedly, Experiment 2 shows that adaptation to the end-effector irrelevant stick- tilt perturbation was accompanied by an increase in the end-effector relevant tip- movement direction error (Fig. 7c). This counterintuitive result can be explained by the constraint of the adaptation pattern along the baseline TMD–STA relationship (Fig. 3a). According to this constraint, CCW (or CW) stick-tilt adaptation should change the tip- movement direction in the CCW (or CW) direction (Fig. 7d). The influence of stick-tilt perturbation on tip-movement direction error tended to be more pronounced for participants with a smaller stick-tilt angle at baseline, i.e., those with a shallower baseline TMD–STA relationship (Fig. 7f), further supporting this idea. The results of Experiment 3 also support the existence of such an interaction between tip-movement and stick-tilt adaptation. When the tip-movement and stick-tilt perturbations were imposed simultaneously, the adaptation trajectories on the physical plane were significantly influenced by the perturbation combination patterns (Fig. 9c). If the motor system had simply attempted to correct the stick-tilt to reduce the sensory prediction error, this spillover effect on the tip-movement direction would not have been observed.

Given that the baseline TMD–STA relationship (Fig. 3a) reflects the motor cost minimization, it is likely that the motor system is attempting to reduce both the sensor prediction error and the motor cost. One remaining question is why the motor system needs to correct for end-effector irrelevant stick-tilt errors in the first place. We speculate that there may be two reasons. First, a possible simple explanation is that the motor system predicts all sensory consequences generated by the action and attempts to correct any differences between the predicted and actual sensory information, even when the task is successfully achieved (Mazzoni & Krakauer, 2006; Schaefer et al., 2012; Morehead et al., 2017; Kim et al., 2019). Similarly, it is possible that the correction for end-effector irrelevant stick errors reflects the motor system’s attempt to resolve the dissociation between stick (i.e., vision) and hands (i.e., proprioception) caused by the perturbation.

Second, given that tip-movement direction and stick-tilt angle are tightly coupled (i.e., the presence of baseline TMD–STA relationship), the presence of an end-effector irrelevant stick-tilt visual error implies the presence of an end-effector relevant tip-movement direction error. In other words, the motor system interprets end- effector irrelevant errors as evidence that end-effector relevant errors have been also produced. When sensory information contains noise and/or uncertainty, the reliability of sensory information estimation can be improved by utilizing additional sensory information (Ernst & Banks, 2002). From this viewpoint, utilizing stick-tilt information may be useful for improving the accuracy of the estimation of the tip-movement direction.

Figure 10 illustrates the possible mechanism by which the motor system alters the movement pattern of the redundant system when encountering visual errors. The motor system develops a stereotyped movement pattern through an optimization process that is expressed by a baseline TMD–STA relationship (gray lines in Figs. 10a and 10b). This relationship not only dictates how the motor system physically controls the redundant system, but also informs the resulting visual feedback. When visual perturbations are applied to the tip-movement direction or the stick-tilt angle (bold arrows in Fig. 10a), the motor system tries to eliminate the dissociation of the perturbed state from the original state (thin arrows in Fig. 10a). It even responds to the end- effector irrelevant perturbation (red and blue arrows in Fig. 10a), because the error in the end-effector irrelevant dimension implies the presence of an end-effector relevant error (black dashed arrows in Fig. 10a) under the condition that the stick-tilt angle is tightly coupled with the tip-movement direction (i.e., the baseline TMD–STA relationship). However, these correction patterns are not physically allowed. Since the tip-movement direction and the stick-tilt angle are physically constrained by the baseline TMD–STA relationship, these correction patterns are biased toward the relationship (bold arrows in Fig. 10b).

A hypothesis for how the motor system alters the movement pattern of redundant system.

The motor system develops a stereotypical relationship in redundant space, possibly through an optimization process (a gray line). (a) In the visual space, this relationship plays a crucial role for the motor system in how it should implicitly respond to the various visual perturbations (bold arrows) in the redundant space. The motor system tries to eliminate the dissociation of the perturbed state from the original state (thin arrows). According to the relationship, the motor system interprets end-effector irrelevant errors (red or blue) as evidence that end-effector relevant errors have been also produced (dashed arrows). (b) However, the correction patterns are constrained by the relationship in the physical space. The relationship determined which movement pattern the motor system should perform. Thus, the correction patterns gravitate toward the relationship (bold arrows).

In summary, this study shows that the baseline TMD–STA relationship, originally developed to control the redundant system, also plays a crucial role for the motor system in how it should implicitly respond to the visual errors and how it should correct the physical movement. This idea provides a novel framework for understanding how the motor system leans to control the redundant system when encountering visual errors.

Materials and methods

Participants

A total of 72 healthy adults (25 females and 47 males, aged 22.1 ± 2.5 years old; mean ± SD) participated in the study. Participants were recruited using an online recruitment system (https://www.jikken-baito.com). All the participants were right- handed according to their laterality score (83.3 ± 13.7; mean ± SD) obtained using the Edinburgh Handedness Inventory (Oldfield, 1971). They provided written informed consent, and the experimental protocol was approved by the ethics committee of The University of Tokyo (#19-225).

Experimental apparatus

The participants performed reaching movements using a virtual stick displayed on a horizontal screen with both handles of a manipulandum (KINARM End-Point Lab; Kinarm, Kingston, Canada) (Scott, 1999) (Fig. 2a). This screen prevented the participants from directly seeing their arms and handles, and displayed a start position, a target (1.4 cm in diameter), and a virtual stick (length: 40 cm, width: 0.5 cm). A white circle (1 cm diameter) was displayed on the right tip of the stick. The participants held the left tip of the stick with their left hand (left handle) and the stick 15 cm away from the left hand with their right hand (right handle). The distance between the handles was fixed using a strong elastic force produced by the manipulandum (spring constant = 2000 N/m). The positions and velocities of the hands were analogue/digitally converted at 1.129 kHz and then recorded at 1 kHz for offline analysis.

Baseline phase

All the experiments (Experiments 1–3) comprised a baseline phase (360 trials) followed by an adaptation phase (240 trials). The baseline phase, which was common to all experiments, was performed to investigate the basic strategy of how the participants moved the stick to reach the targets with the right tip. Before each reaching movement, the participants were asked to move the stick to the home position with the help of a manipulandum (Fig. 2c). After maintaining the home position for 0.5−1 s, a gray target was pseudo-randomly chosen from nine possible targets (0°: horizontal direction, ±10°, ±20°, ±30°, ±40°) appearing 10 cm away from the initial right tip position of the stick. After waiting for another 0.5−1 s, the color of the target changed to pink, indicating that they should immediately start moving the tip to the target. The participants were instructed to maintain the peak velocities of the tip as constant as possible across the trials. When the tip reached the target, a feedback message, “fast” or “slow,” was presented on the monitor if the peak movement speed was higher or lower than the range of 300−450 mm/s, respectively. After maintaining the tip at the target for 0.5 s, the visual feedback of the stick disappeared. Then, the handles automatically returned to their home position, and the next trial began. During the subsequent 240 trials in the adaptation phase, the target appeared only in the 0° direction, and distinct visual perturbations were introduced in Experiments 1–3 as described below. The home positions, size of the visual feedback and the target, inter- trial interval, and criteria for the tip velocity feedback for the adaptation phase were the same as those in the baseline phase.

Experiment 1

Experiment 1A (N = 19) and Experiment 1B (N = 13) were performed to investigate how participants moved the stick when adapting to end-effector relevant perturbations (tip perturbations). In Experiment 1A, the tip-movement direction on the monitor was gradually rotated CCW direction around the starting position by 1° per trial up to 30° (Fig. 4a). The stick-tilt angle remained unchanged (i.e., the stick-tilt angle was identical to the angle of the line connecting both hands). In Experiment 1A, participants did not perceive the presence of rotation and the adaptation was considered to be dominated by the implicit process. On the other hand, in Experiment 1B, the 30° visual rotation was abruptly applied to the tip-movement direction, which allowed them to use an explicit strategy.

Experiment 2

Experiment 2 (N = 20) examined whether the motor system was indifferent to end-effector irrelevant perturbations (stick-tilt perturbations). The stick-tilt angle on the monitor was rotated around the right tip in the CW direction (E2CW group, N = 10) or in the CCW direction (E2CCW group, N = 10). Notably, these perturbations did not affect the tip-movement direction. During each reaching trial, the stick-tilt angle increased from 0° to 6° in proportion to the distance of the tip position from the starting point (0.6°/cm) (Fig. 6a). Participants did not notice any perturbations during the experiment.

Experiment 3

In Experiment 3 (N = 20), to examine how end-effector irrelevant stick-tilt perturbations influenced adaptation to end-effector relevant tip perturbations, both tip-movement and stick-tilt perturbations were simultaneously imposed (Fig. 8a). Ten participants participated in a task in which CCW tip and CCW stick-tilt rotations were imposed (E3CCW group). The remaining ten participants participated in a task in which CCW tip and CW stick-tilt rotations were imposed (E3CW group). The procedure for applying tip and stick-tilt perturbations was the same as in Experiment 1A and Experiment 2, respectively. The participants did not notice perturbations during the experiment.

Data analysis

Data analyses were performed using MATLAB R2023a (MathWorks, Natick, Massachusetts, USA). The behavioral data were low-pass filtered (10 Hz, fourth-order Butterworth filter). The tip-movement direction was calculated as the movement direction of the tip relative to the horizontal direction (Fig. 3a). The stick-tilt angle was defined as the tilting angle of the stick from the horizontal line (Fig. 3a). Positive values for both parameters were defined as the degree to the CCW direction. Unless otherwise noted, the tip-movement direction and the stick-tilt angle at the tip’s peak velocity were used to evaluate the feedforward motor commands.

Minimum distance of hand movement

The total movement distance of both hands in Figs. 3b and 3c was estimated by summing the Euclidean distances between the initial and final positions of the hands. The distance is represented as a function of the stick-tilt angle at the movement offset (θ). The first term on the right side represents the distance of the left-hand movement, whereas the second term represents the distance of the right-hand movement. ϕ, d, L, and R represent a target direction (0°, ±10°, ±20°, ±30°, and ±40°), a target distance (10 cm), a distance between the left hand and the tip (40 cm), and a distance between the right hand and the tip (25 cm), respectively. The minimum distance of hand movement was derived by solving the following equation:

Statistical analysis

To statistically evaluate the behavioral data, we conducted t-tests, correlation analyses, two-way repeated-measures ANOVA, and MANOVA using MATLAB, the software package JASP (http://jasp-stats.org), and R studio.

Data Availability

Data collected in Experiments 1–3 and the code are uploaded in zenodo and are available at https://zenodo.org/record/8033232.